---
title: "Flaw of Averages: the Risk FEMA’s National Risk Index Doesn’t Report"
author: "Emily Huang"
format: 
  html: 
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    code-fold: true
    code-summary: "Show Code"
    theme: cosmo
    fig-width: 10
    fig-height: 5
    df-print: paged
    embed-resources: true
editor: 
  markdown: 
    wrap: 70
---

# NRI Reconstruction

## Load Packages

```{r initial_setup}
#| label: initial setup

# Loading packages 
library(tidyverse)
library(conflicted) # To resikve name conflicts between packages
library(scales)
library(performance) # for over_dispersion()
library(MASS) # Fit distribution for negative Bernoulli MLE
library(patchwork)
library(mvtnorm) # For t-copula
library(fitdistrplus) # To fit historical data to distribution via MLE
#library(copula)
library(nortest) # For Anderson-Darling goodness-of-fit test 
#library(Matrix)
library(knitr) # For ktables, which makes nice tables
library(kableExtra) # For more table creation and styling abilities
library(pscl) # For functions to handle zero-inflated Poisson model

# Resolve conflict between packages
conflict_prefer("select",    "dplyr")
conflict_prefer("filter",    "dplyr")
conflict_prefer("summarise", "dplyr")
conflict_prefer("rename",    "dplyr")
conflict_prefer("lag",       "dplyr")
```

## Purpose

This document contains all the R code for my senior project. In
particular, it includes 1) the reconstruction of FEMA's National Risk
Index (NRI) for New York County (Manhattan, FIPS 36061) using FEMA's
published technical documentation and loss data from SHELDUS and 2)
improving FEMA's point-estimate measure of risk by replacing 3 of the
NRI input variables with fitted probability distributions and
recalculate the risk with a Monte Carlo Simulation of 100,000 random
draws. The goal is to make a point that FEMA's report of a single
mean-based risk value is a gross oversimplification of the actual risk
experienced by communities, especially when the probability
distribution of variables that contribute to the risk calculation
often have fat tails.

**Note:** Within the period that we are looking at, New York County
(Manhattan) has zero recorded riverine flooding building losses in the
SHELDUS data set (1960-2022) because losses from hurricane Sandy was
recorded under "Hurricane/Tropical Storm" instead. Thus, I was not
able to fit a distribution through MLE with 0 observations. FEMA
accounted for this through their own proprietory processing, which I
wasn't able to replicate. So for the Monte Carlo Simulation, we used
RFDL data from a larger geographical scale, NY state regional scale
LRB, by pooling loss ratio observations from all counties across NYS
that did have recorded riverine flood losses to fit the distribution.
This fallback likely made the RFLD component conservative by
understating the true risk experienced by members of the New York
County. Underestimation is likely the case because Manhattan has the
higher building value per square mile than any other counties in New
York State, so when a flood hits, the dollar loss amount should be
much higher relative to most NYS counties.

## Load Data

Two data sources are loaded here, 1) SHELDUS historical loss records
and 2) FEMA's published National Risk Index and their input variables.

1.  **SHELDUS v24** (Feburary 2026): https://cemhs.asu.edu/sheldus

2.  **FEMA NRI v1.20** (December 2025):
    https://www.fema.gov/flood-maps/products-tools/national-risk-index

**Note:** FEMA's NRI v1.20 selected specific date ranges for the
SHELDUS data, which is unbeknownst to me. To resokve this, I used
min/max year for each hazard in the county from the SHELDUS data
instead, which likely is the cause of discrepencies in Annualized
Frequency component of the reconstructed NRI.

```{r load_data}
sheldus_raw <- read_csv(
  "C:/Users/emily/Desktop/Econ-Fin Senior Project/FEMA NRI/data/Other/SHELDUS.csv",
  show_col_types = FALSE
) %>%
  rename(
    prop_dmg_adj = `PropertyDmg(ADJ 2022)`,
    crop_dmg_adj = `CropDmg(ADJ 2022)`
  )

nri <- read_csv(
  "C:/Users/emily/Desktop/Econ-Fin Senior Project/FEMA NRI/data/NRI_Table_Counties.csv",
  show_col_types = FALSE
)

ny <- nri %>% filter(STCOFIPS == "36061") # ny is for new york county only (manhattan)

```

## Geographical Scales

FEMA's Bayesian credibility analysis pools Loss Ratio Basis (LRB)
across 4 nested geographical levels (county, surrounding, regional,
and national) to update estimations of risk for data-sparse counties
by blending data with broader geographical scale.

```{r scales}
#| label: scales

# County: New York County only
fips_county <- 36061

# Surrounding: the five NYC boroughs plus adjacent NJ metro counties
fips_surrounding <- c(    
  36005,  # Bronx
  36047,  # Kings (Brooklyn)
  36081,  # Queens
  36085,  # Richmond (Staten Island)
  34017,  # Hudson, NJ
  34003,  # Bergen, NJ
  34013,  # Essex, NJ
  36119   # Westchester, NY
)

# Regional: all New York State counties (FIPS 36***)
fips_regional <- nri %>%
  filter(str_starts(as.character(STCOFIPS), "36")) %>%
  pull(STCOFIPS)

# National: all counties in SHELDUS
fips_national <- unique(sheldus_raw$County_FIPS)

# Extract SHELDUS subsets for each scale
sheldus_county   <- sheldus_raw %>% filter(County_FIPS == fips_county)
sheldus_surround <- sheldus_raw %>% filter(County_FIPS %in% fips_surrounding)
sheldus_regional <- sheldus_raw %>% filter(County_FIPS %in% fips_regional)
sheldus_national <- sheldus_raw


# SHELDUS record counts at each geographic scale
tibble(
  Scale      = c("County", "Surrounding", "Regional (NY State)", "National"),
  n_counties = c(1, length(fips_surrounding),
                 length(fips_regional), length(fips_national)),
  n_rows     = c(nrow(sheldus_county), nrow(sheldus_surround),
                 nrow(sheldus_regional), nrow(sheldus_national))
) %>%
  kable(caption = "SHELDUS record counts at each geographic scale") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

## Hazard Mapping

We map SHELDUS's free-text label to NRI's hazard names and define
occurrence basis using FEMA's Hazard Occurrence basis chart. There are
only 12 active NRI hazard types applicable to New York County. Hazards
ommitted from Manhattan includes Avalanche, Volcano, Tsunami,
Wildfire, Drought and Landslide.

```{r hazards}
# SHELDUS label -> NRI prefix and occurrence basis
#   "event"      = distinct events (count rows)
#   "event_day"  = event-days (sum Duration_Days)
hazard_map <- tribble(
  ~sheldus_label,               ~nri_prefix, ~occurrence_basis,
  "Flooding",                   "RFLD",      "event_day",
  "Coastal",                    "CFLD",      "none",
  "Hurricane/Tropical Storm",   "HRCN",      "event",
  "Severe Storm/Thunder Storm", "SWND",      "event",
  "Wind",                       "SWND",      "event",
  "Winter Weather",             "WNTW",      "event_day",
  "Hail",                       "HAIL",      "event",
  "Lightning",                  "LTNG",      "event",
  "Heat",                       "HWAV",      "event_day",
  "Tornado",                    "TRND",      "event",
  "Fog",                        "SWND",      "event"
)

active_hazards <- c("CFLD","RFLD","HRCN","ERQK","SWND","TRND",
                    "HAIL","LTNG","HWAV","CWAV","WNTW","ISTM")

hazard_map %>%
  kable(caption = "SHELDUS label → NRI prefix mapping") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

## SHELDUS Pre-Processing

According to FEMA's technical documentation, there is a 3-step
pre-processing procedure to prepare SHELDUS records for FEMA's NRI
calculations:

1.  **Record expansion:** divides multi-day event losses equally
    across days. We do this because event-day hazards (like RFLD,
    WNTW, or HWAV) count Duration_Days as the frequency denominator;
    without expansion, all loss from a 5-day flood would be attributed
    to a single day-count, which might lead to overestimating.

2.  **Record aggregation:** merges records with the same county, year,
    month, and hazard into one row, preventing double-counting
    events/event days

3.  **Zero-loss insertion:** when historical occurrence data indicates
    more events than loss-causing records, zero-loss records are
    inserted to avoid overestimating average impact. This is done
    already because instead of inserting lrb=0 for all row without
    occurrences, the formula used the whole period of record as
    denominator. This essentially makes sure we consider this.

```{r pre-processing}
event_day_labels <- c("Flooding", "Winter Weather", "Heat")

expand_to_daily <- function(df) {   # function for record expansion (step 1)
  event_rows     <- df %>% filter(Hazard %in% event_day_labels, Duration_Days > 1)
  non_event_rows <- df %>% filter(!(Hazard %in% event_day_labels & Duration_Days > 1))
  expanded <- event_rows %>%
    uncount(Duration_Days, .id = "day_id", .remove = FALSE) %>%
    mutate(
      prop_dmg_adj  = prop_dmg_adj  / Duration_Days,
      crop_dmg_adj  = crop_dmg_adj  / Duration_Days,
      Fatalities    = Fatalities    / Duration_Days,
      Injuries      = Injuries      / Duration_Days,
      Duration_Days = 1L
    )
  bind_rows(non_event_rows, expanded)
}

# perform step 1 to all four geographical levels
sheldus_expanded_county   <- expand_to_daily(sheldus_county)
sheldus_expanded_surround <- expand_to_daily(sheldus_surround)
sheldus_expanded_regional <- expand_to_daily(sheldus_regional)
sheldus_expanded_national <- expand_to_daily(sheldus_national)

aggregate_records <- function(df) {   # function for aggregating records (step 2)
  df %>%
    group_by(County_FIPS, Hazard, Year, Month) %>%
    summarise(
      prop_dmg_adj  = sum(prop_dmg_adj,  na.rm = TRUE),
      crop_dmg_adj  = sum(crop_dmg_adj,  na.rm = TRUE),
      Fatalities    = sum(Fatalities,    na.rm = TRUE),
      Injuries      = sum(Injuries,      na.rm = TRUE),
      Duration_Days = sum(Duration_Days, na.rm = TRUE),
      n_records     = n(),
      .groups       = "drop"
    )
}

# perform step 2 to all four geographical levels
sheldus_county_clean   <- aggregate_records(sheldus_expanded_county)
sheldus_surround_clean <- aggregate_records(sheldus_expanded_surround)
sheldus_regional_clean <- aggregate_records(sheldus_expanded_regional)
sheldus_national_clean <- aggregate_records(sheldus_expanded_national)


# create summary table and to check the total property damage remains the same before and after the data processing
tibble(
  Step            = c("Raw", "After expansion", "After aggregation"),
  Total_Prop_Loss = c(
    sum(sheldus_county$prop_dmg_adj,          na.rm = TRUE),
    sum(sheldus_expanded_county$prop_dmg_adj, na.rm = TRUE),
    sum(sheldus_county_clean$prop_dmg_adj,    na.rm = TRUE)
  )
) %>%
  mutate(Total_Prop_Loss = dollar(round(Total_Prop_Loss))) %>%
  kable(caption = "Loss preservation check: total property damage through pre-processing") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

## Exposure

Exposure is the dollar value of the assets under risk due to natural
hazards. We use FEMA's pre-computed NRI table directly because the
Hazus/CoreLogic data is proprietary. If we use raw data to calculate
on our own, there will be a large discrepancy and room for error when
we compare our reconstructed NRI with FEMA's due to lack of data
transparency. The population column (`_EXPP`) is a person count; we
convert to dollars using FEMA's implied VSL of \$13,702,913, which is
back-calculated from Cold Wave totals.

```{r exposure}
#| label: exposure

FEMA_VSL <- 13702913

# Calculating exposure for all active hazards
exposure_ny <- map_dfr(active_hazards, function(haz) {
  expb_col <- paste0(haz, "_EXPB")
  expp_col <- paste0(haz, "_EXPP")
  expa_col <- paste0(haz, "_EXPA")
  expb <- if (expb_col %in% names(ny)) replace_na(ny[[expb_col]], 0) else 0
  expp <- if (expp_col %in% names(ny)) replace_na(ny[[expp_col]], 0) * FEMA_VSL else 0
  expa <- if (expa_col %in% names(ny)) replace_na(ny[[expa_col]], 0) else 0
  tibble(
    nri_prefix       = haz,
    consequence      = c("Building", "Population", "Agriculture"),
    exposure_density = c(expb, expp, expa)
  )
}) %>%
  filter(exposure_density > 0)

# Verification for CWAV because it is the only hazard that lets me confirm the VSL conversion factor is right.
ny %>%
  select(CWAV_EXPB, CWAV_EXPP, CWAV_EXPA, CWAV_EXPT) %>%
  mutate(
    recon_expt = CWAV_EXPB + CWAV_EXPP * FEMA_VSL + replace_na(CWAV_EXPA, 0),
    ratio      = round(recon_expt / CWAV_EXPT, 4)
  ) %>%
  kable(caption = "VSL verification: ratio should be 1.00") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

## Annualized Frequency

The annualized frequency is calculated by $AF$ = occurrence count ÷
period of record. This is part of the 3 multiplicative terms that we
need to get the Estimated Annual Loss estimation for Manhattan
$EAL = \text{Exposure} \times AF \times HLR$.

**Note:** Notice from the formula that value for annualized frequency
depends on the period of record, but FEMA does not publish the exact
window in which they extracted SHELDUS data from. We use the min/max
year to find period of record in our SHELDUS dataset to determine PoR
(period of Record) for each hazard. This mismatch results in different
set of hazard, and therefore their losses, to be considered in our
reconstruction. Thus, there are discrepancies in our risk estimate
when compared to FEMA's published estimates.

```{r frequency}
#| label: frequency

empirical_por <- sheldus_county_clean %>%
  group_by(Hazard) %>%
  summarise(
    min_year      = min(Year),
    max_year      = max(Year),
    por_empirical = max(Year) - min(Year) + 1,
    .groups       = "drop"
  )

por_override <- tribble(
  ~sheldus_label,               ~por_fema,
  "Flooding",                    NA_real_,
  "Hurricane/Tropical Storm",    NA_real_,
  "Winter Weather",              NA_real_,
  "Severe Storm/Thunder Storm",  NA_real_,
  "Wind",                        NA_real_,
  "Heat",                        NA_real_,
  "Hail",                        NA_real_,
  "Lightning",                   NA_real_,
  "Tornado",                     NA_real_
)

por_table <- empirical_por %>%
  left_join(por_override, by = c("Hazard" = "sheldus_label")) %>%
  mutate(por_used = if_else(!is.na(por_fema), por_fema, por_empirical))

compute_af_county <- function(df_clean, por_tbl, hazard_label, basis) {
  sub <- df_clean  %>%  filter(County_FIPS == fips_county, Hazard == hazard_label)
  por <- por_tbl %>% filter(Hazard == hazard_label) %>% pull(por_used)
  if (length(por) == 0 || is.na(por)) por <- 1
  occurrence_count <- if (basis == "event_day") {
    sum(sub$Duration_Days, na.rm = TRUE)
  } else {
    nrow(sub)
  }
  tibble(
    sheldus_label    = hazard_label,
    basis            = basis,
    occurrence_count = occurrence_count,
    por_years        = por,
    af               = occurrence_count / por
  )
}

af_inputs <- hazard_map %>%
  select(sheldus_label, nri_prefix, occurrence_basis) %>%
  filter(occurrence_basis != "none") %>%
  distinct(sheldus_label, .keep_all = TRUE)

af_county <- pmap_dfr(
  list(af_inputs$sheldus_label, af_inputs$occurrence_basis),
  function(lbl, basis) compute_af_county(sheldus_county_clean, por_table, lbl, basis)
) %>%
  left_join(af_inputs, by = "sheldus_label")


af_county_dedup <- af_county %>%
  group_by(nri_prefix) %>%
  slice_max(af, n = 1, with_ties = FALSE) %>%
  ungroup()

# FEMA published AF for comparison
af_fema_lookup <- tibble(
  nri_prefix = unique(af_inputs$nri_prefix),
  af_fema    = map_dbl(unique(af_inputs$nri_prefix), ~ {
    col <- paste0(.x, "_AFREQ")
    if (col %in% names(ny)) ny[[col]] else NA_real_
  })
)

# Compare to FEMA published annualized frequency where available
af_compare <- af_county_dedup %>%
  left_join(af_fema_lookup, by = "nri_prefix") %>%
  mutate(ratio_to_fema = round(af / af_fema, 3))

af_compare %>%
  select(sheldus_label, nri_prefix, basis, occurrence_count,
         por_years, af, af_fema, ratio_to_fema) %>%
  kable(digits = c(0,0,0,0,0,3,3,3),
        caption = "Reconstructed vs. FEMA annualized frequency") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

## Historic Loss Ratio (HLR)

THE HLR is the average of exposed value that is lost per hazard
occurrence. FEMA estimated using Bayesian Credibility weighting across
4 of the geographical scales.

$$LRB_{dhgc} = \frac{\text{Loss}_{dhgc}}{\text{Exposure}_{hgc}} \qquad
  W_{hgc} = \frac{1/\text{var}_{hgc}}{\sum_g 1/\text{var}_{hgc}} \qquad
  HLR_{hc} = \sum_g \overline{LRB}_{hgc} \times W_{hgc}$$

### HLR Exposure Data

```{r hlr-exposure}
#| label: hlr-exposure

hlr_exposure_all_counties <- map_dfr(active_hazards, function(haz) {
  expb_col <- paste0(haz, "_EXPB")
  expp_col <- paste0(haz, "_EXPP")
  expa_col <- paste0(haz, "_EXPA")
  nri %>%
    select(STCOFIPS, any_of(c(expb_col, expp_col, expa_col))) %>%
    mutate(
      nri_prefix = haz,
      hlr_exp_b  = if (expb_col %in% names(nri)) replace_na(.data[[expb_col]], 0) else 0,
      hlr_exp_p  = if (expp_col %in% names(nri)) replace_na(.data[[expp_col]], 0) * FEMA_VSL else 0,
      hlr_exp_a  = if (expa_col %in% names(nri)) replace_na(.data[[expa_col]], 0) else 0
    ) %>%
    select(STCOFIPS, nri_prefix, hlr_exp_b, hlr_exp_p, hlr_exp_a) %>%
    pivot_longer(
      cols      = c(hlr_exp_b, hlr_exp_p, hlr_exp_a),
      names_to  = "consequence",
      values_to = "hlr_exposure"
    ) %>%
    mutate(consequence = case_when(
      consequence == "hlr_exp_b" ~ "Building",
      consequence == "hlr_exp_p" ~ "Population",
      consequence == "hlr_exp_a" ~ "Agriculture"
    ))
}) %>%
  filter(hlr_exposure > 0)

hlr_exposure_all_counties
```

### Compute LRB

For each geographical scale.

```{r lrb}
#| label: lrb

# VSL value directly from the technical documentation
VSL_fatality <- 11.6e6
VSL_injury   <- VSL_fatality / 10

# function to compute loss ratio basis
compute_lrbs <- function(df_clean, scale_label, hlr_exp_all) {
  df_clean %>%
    left_join(hazard_map, by = c("Hazard" = "sheldus_label")) %>%
    filter(!is.na(nri_prefix)) %>%
    mutate(
      loss_building    = replace_na(prop_dmg_adj, 0),
      loss_population  = replace_na(Fatalities, 0) * VSL_fatality +
                         replace_na(Injuries,   0) * VSL_injury,
      loss_agriculture = replace_na(crop_dmg_adj, 0)
    ) %>%
    pivot_longer(
      cols      = c(loss_building, loss_population, loss_agriculture),
      names_to  = "loss_type",
      values_to = "loss"
    ) %>%
    mutate(consequence = case_when(
      loss_type == "loss_building"    ~ "Building",
      loss_type == "loss_population"  ~ "Population",
      loss_type == "loss_agriculture" ~ "Agriculture"
    )) %>%
    left_join(
      hlr_exp_all %>% rename(County_FIPS = STCOFIPS),
      by = c("County_FIPS", "nri_prefix", "consequence")
    ) %>%
    filter(!is.na(hlr_exposure), hlr_exposure > 0) %>%
    mutate(lrb = loss / hlr_exposure, scale = scale_label) %>%
    select(nri_prefix, consequence, County_FIPS, Year, Month, lrb, scale)
}

# compute lrb for all 4 of the geographical levels
lrb_county   <- compute_lrbs(sheldus_county_clean,   "county",      hlr_exposure_all_counties)
lrb_surround <- compute_lrbs(sheldus_surround_clean, "surrounding", hlr_exposure_all_counties)
lrb_regional <- compute_lrbs(sheldus_regional_clean, "regional",    hlr_exposure_all_counties)
lrb_national <- compute_lrbs(sheldus_national_clean, "national",    hlr_exposure_all_counties)

lrb_all <- bind_rows(lrb_county, lrb_surround, lrb_regional, lrb_national)

lrb_all %>% count(scale) %>%
  kable(caption = "LRB observation counts by geographic scale") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

### Bayesian Credibility Weighting

```{r bayesian-hlr}
#| label: bayesian-hlr

# Following formula to reconstruct the HLR component of NRI, then comparing it to FEMA's
lrb_stats <- lrb_all %>%
  group_by(nri_prefix, consequence, scale) %>%
  summarise(
    n_events = n(),
    avg_lrb  = mean(lrb, na.rm = TRUE),
    var_lrb  = var(lrb,  na.rm = TRUE),
    .groups  = "drop"
  ) %>%
  mutate(var_lrb = if_else(is.na(var_lrb) | var_lrb == 0, 1e-20, var_lrb))

lrb_weights <- lrb_stats %>%
  group_by(nri_prefix, consequence) %>%
  mutate(inv_var = 1 / var_lrb, weight = inv_var / sum(inv_var)) %>%
  ungroup()

hlr_bayesian <- lrb_weights %>%
  group_by(nri_prefix, consequence) %>%
  summarise(HLR_bayesian = sum(avg_lrb * weight), .groups = "drop")

hlr_fema_cols <- c(outer(active_hazards, c("_HLRB","_HLRP","_HLRA"), paste0))

hlr_fema <- ny %>%
  select(any_of(hlr_fema_cols)) %>%
  pivot_longer(everything(), names_to = "col", values_to = "HLR_fema") %>%
  mutate(
    nri_prefix  = str_sub(col, 1, 4),
    consequence = case_when(
      str_ends(col, "_HLRB") ~ "Building",
      str_ends(col, "_HLRP") ~ "Population",
      str_ends(col, "_HLRA") ~ "Agriculture"
    )
  ) %>%
  select(nri_prefix, consequence, HLR_fema)

hlr_bayesian %>%
  left_join(hlr_fema, by = c("nri_prefix","consequence")) %>%
  mutate(
    ratio        = round(HLR_bayesian / HLR_fema, 3),
    HLR_bayesian = signif(HLR_bayesian, 5),
    HLR_fema     = signif(HLR_fema, 5)
  ) %>%
  filter(!is.na(HLR_fema)) %>%
  arrange(nri_prefix, consequence) %>%
  kable(caption = "Reconstructed vs. FEMA published HLR by hazard-consequence") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

## LRB distribution summary

We analyze the Coefficient of Variation (CV=SD/MEAN) of the Loss Ratio
Basis distribution to help us understand how poorly FEMA's mean-based
HLR point estimate represents the real variability of loss. For
hazards with CV\>1, we know the mean is dominated by rare and extreme
events, which are precisely the ones misrepresented by a mean-based
point estimate.

```{r cv-analysis}
#| label: cv-analysis
#| fig-height: 6

cv_summary <- lrb_all %>%
  filter(scale == "county") %>%
  group_by(nri_prefix, consequence) %>%
  summarise(
    n  = n(),
    mean = mean(lrb, na.rm = TRUE),
    sd   = sd(lrb,   na.rm = TRUE),
    cv   = sd / mean,
    .groups = "drop"
  )

cv_plot <- cv_summary %>%
  filter(consequence == "Building") %>%
  mutate(nri_prefix = reorder(nri_prefix, cv))

ggplot(cv_plot, aes(x = nri_prefix, y = cv, fill = cv > 1)) +
  geom_col(alpha = 0.85, width = 0.7) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "firebrick", linewidth = 0.9) +
  scale_fill_manual(
    values = c("FALSE" = "steelblue", "TRUE" = "firebrick"),
    labels = c("FALSE" = "CV ≤ 1", "TRUE" = "CV > 1 (priority for MC)"),
    name   = NULL
  ) +
  annotate("text", x = 0.6, y = 1.05, label = "CV = 1 threshold",
           color = "firebrick", hjust = 0, size = 3.2) +
  coord_flip() +
  labs(
    title    = "Coefficient of Variation of LRBs by Hazard — Building, County Scale",
    x = "NRI Hazard Prefix", y = "CV (SD / Mean)"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"), legend.position = "bottom")
```

## Expected Annual Loss (EAL)

After calculating the Exposure, Annualized Frequency, and Historic
Loss Ratio, we assemble all three components to compute the Expected
Annual Loss. Then we compare it to FEMA's published value.

```{r eal}
#| label: eal

# Use FEMA's published AF for all hazards. This avoids contaminating the EAL
af_full <- tibble(
  nri_prefix = c("ERQK","CWAV","CFLD","ISTM","HWAV","SWND","HRCN",
                 "HAIL","TRND","LTNG","WNTW","RFLD"),
  af = map_dbl(c("ERQK","CWAV","CFLD","ISTM","HWAV","SWND","HRCN",
                 "HAIL","TRND","LTNG","WNTW","RFLD"), ~ {
    col <- paste0(.x, "_AFREQ")
    if (col %in% names(ny)) replace_na(ny[[col]], 0) else 0
  })
)

hlr_full <- bind_rows(
  hlr_bayesian %>% rename(HLR = HLR_bayesian),
  hlr_fema     %>% rename(HLR = HLR_fema) %>%
    anti_join(hlr_bayesian, by = c("nri_prefix","consequence"))
)

eal_by_hc <- exposure_ny %>%
  left_join(af_full,  by = "nri_prefix") %>%
  left_join(hlr_full, by = c("nri_prefix","consequence")) %>%
  mutate(
    af               = replace_na(af, 0),
    HLR              = replace_na(HLR, 0),
    exposure_density = replace_na(exposure_density, 0),
    EAL_hc           = exposure_density * af * HLR
  ) %>%
  select(nri_prefix, consequence, exposure_density, af, HLR, EAL_hc) %>%
  arrange(nri_prefix, consequence)

eal_by_hc %>%
  mutate(
    exposure_density = dollar(round(exposure_density)),
    af               = round(af, 4),
    HLR              = signif(HLR, 4),
    EAL_hc           = dollar(round(EAL_hc))
  ) %>%
  kable(caption = "EAL per hazard-consequence: EAL = Exposure × AF × HLR") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)

eal_by_hazard <- eal_by_hc %>%
  group_by(nri_prefix) %>%
  summarise(EAL_h_calc = sum(EAL_hc, na.rm = TRUE), .groups = "drop")

eal_fema_by_hazard <- ny %>%
  select(any_of(paste0(active_hazards, "_EALT"))) %>%
  pivot_longer(everything(), names_to = "col", values_to = "EAL_fema") %>%
  mutate(nri_prefix = str_sub(col, 1, 4)) %>%
  select(nri_prefix, EAL_fema)

eal_by_hazard %>%
  left_join(eal_fema_by_hazard, by = "nri_prefix") %>%
  mutate(
    ratio      = round(EAL_h_calc / replace_na(EAL_fema, 1), 3),
    EAL_h_calc = dollar(round(EAL_h_calc)),
    EAL_fema   = dollar(round(replace_na(EAL_fema, 0)))
  ) %>%
  arrange(desc(as.numeric(gsub("[^0-9.]", "", EAL_h_calc)))) %>%
  kable(col.names = c("Hazard","Reconstructed EAL","FEMA EAL","Ratio"),
        caption   = "EAL by hazard: Ratio = Reconstructed / FEMA.") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)

composite_eal_calc <- sum(eal_by_hc$EAL_hc, na.rm = TRUE)
composite_eal_fema <- ny$EAL_VALT


# comparing our reconstructed EAL with FEMA's
tibble(
  Source        = c("Reconstructed", "FEMA Published"),
  Composite_EAL = dollar(round(c(composite_eal_calc, composite_eal_fema))),
  Pct_of_FEMA   = round(c(composite_eal_calc / composite_eal_fema * 100, 100), 1)
) %>%
  kable(caption = "Composite EAL comparison") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

## Community Risk Factor

With the Expected Annual Loss computed, the last step to get a risk
estimate for Manhattan is accounting for their unique Community Risk
Factor by $$\text{Risk Value} = EAL \times CRF.$$ The CRF is derived
by ranking the ratio of each county's Social Vulnerability Index (SVI)
and Community Resilience Index (CRI) nationally and mapping through a
triangular CDF (with min=0.5, max=2, and mode=1)

```{r crf-rg}
#| label: crf-rg

crf_data <- nri %>%
  filter(!is.na(SOVI_SCORE), !is.na(RESL_SCORE), RESL_SCORE > 0) %>%
  mutate(r_g = SOVI_SCORE / RESL_SCORE)

r_ny <- crf_data %>% filter(STCOFIPS == "36061") %>% pull(r_g)

cat("New York County r_g (SVI/CRI) =", round(r_ny, 4), "\n")
```

```{r crf-calc}
#| label: crf-calc

crf_data <- crf_data %>% mutate(percentile_r = percent_rank(r_g))
percentile_ny <- crf_data %>% filter(STCOFIPS == "36061") %>% pull(percentile_r)

triangular_inv_cdf <- function(x) {
  case_when(
    x < 0    ~ 0.5,
    x > 1    ~ 2.0,
    x <= 1/3 ~ 0.5 + sqrt(3 * x / 2),
    TRUE     ~ (4 - sqrt(6 - 6 * x)) / 2
  )
}

crf_data    <- crf_data %>% mutate(CRF_calc = triangular_inv_cdf(percentile_r))
crf_ny_calc <- crf_data %>% filter(STCOFIPS == "36061") %>% pull(CRF_calc)
crf_ny_fema <- ny$RISK_VALUE / ny$EAL_VALT


# comparing the reconstructed Community Resilience with FEMA's 
tibble(
  Source     = c("Reconstructed (triangular CDF)", "FEMA Implied (RISK_VALUE/EAL_VALT)"),
  CRF        = round(c(crf_ny_calc, ny$CRF_VALUE), 5),
  SVI_value  = rep(round(ny$SOVI_SCORE, 4), 2),
  CRI_value  = rep(round(ny$RESL_SCORE, 4), 2),
  r_g        = rep(round(r_ny, 4), 2),
  Percentile = rep(round(percentile_ny, 4), 2)
) %>%
  kable(caption = "CRF comparison: reconstructed vs. FEMA implied") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

**Note:** There is a gap between reconstructed CRF = 1.23 vs. FEMA
implied = 1.62. Figure out what was wrong is this...

# Monte Carlo

## Purpose

Each of the following component has a whole probability distribution
of possible values, but FEMA reports a mean-based single point
estimate instead.

$$
\text{EAL} = \text{Exposure} \times \text{Annualized Frequency} \times \text{Historic Loss Ratio}.
$$

This results a single point estimation of risk as well, which results
in the loss of a lot of interesting dimensions of what the risk
estimate could have been. This singular number hides the fact that
hazards cause zero losses in most years and causes catastrophic damage
in a handful of other years. We also will not be able to tell if bad
floods years and bad tornado years tend to happen together. The CV
plot above showed that for most hazards, the mean is dominated by a
few extreme events and are not representative of a typical year.

We wish to illustrate this point more clearly through a Monte Carlo
simulation by asking, what does the full distribution of Expected
Annual Loss should actually look like? We first replace 3 input
variables with a fitted probability distribution based on historical
data. Then instead of collapsing valuable information to a single
mean, we simulate 100,000 hypothetical years, each one drawing from
the fitted probability distributions. At the end of the simulation, we
will have 100,000 loss values, which is an estimate how a full
probability distribution of loss should look like.

The three county level variables we replaced with are 1. Annual
Frequency of Tornado (TRND) 2. Historic Loss Ratio of Flood (RFLD
+CFLD) 3. Historic Loss Ratio of building loss due to Earthquakes
(ERQK). We first (1) fit each of the three variables to a probability
distribution separately using their historic SHELDUS data, this is
called a marginal distribution. Then we (2) simulate all three hazards
together to account for their natural correlation using the t-Copula.

## Part 1: Marginal Distribution and t-Copula

### Underlying theory

If we only replace one of hazard variables, the simulation will only
consist of fitting their distribution to the historical data then
drawing random samples from it. But we are replacing three dependent
variables, so we must simulate them jointly to account for their
natural correlation. For examples, when bad tornado years and bad
flood years tend to occur simultaneously because of severe weather
seasons. Therefore, we must not naively draw random samples from each
distributions independently. We also should not fit them to a single
multivariate distribution to all three because they can have very
different underlying shapes. Thus, we use a copula to solve this by
splitting the probability fitting problem into two because we can
break any joint distribution into two pieces: the shape of each
individual variable (marginal distribution), and how they move
together (their dependence). We will specify each piece independently
and combine them. Formally, by the Skylar's Theorem, for any joint
distribution $F(x_1, x_2, ..,x_n)$ with individual marginal
distributions $F_1, F_2, ...,F_n$, there exists a copula function $C$
such that:

$$
F(x_1, x_2,..., x_n) = C\bigl(F_1(x_1),\, F_2(x_2),\,..., F_n(x_n)\bigr).
$$

### The Gaussian Copula

There are many copulas, but we use the t-Copula, which uses the
correlation structure of a multivariate t-distribution (only the
correlation structure). Notice the marginal distributions do not need
to follow a t-distribution to use this correlation structure. The
t-Copula requires two inputs: the correlation matrix $\mathbf{P}$,
which is a table of pairwise correlations between each of the three
hazards, and a degrees of freedom parameter $\nu$, which controls how
much additional probability is placed on joint tail events. We chose
the t-Copula over the simpler Gaussian Copula because the Gaussian
Copula has asymptotically zero tail dependence (it assigns
increasingly little probability to two hazards being simultaneously
extreme, even when their correlation is nonzero). The t-Copula
corrects this by placing more weight on joint tail outcomes, which is
precisely the behavior we care about when studying catastrophic loss
years. Lower values of $\nu$ produce heavier joint tails, with the
Gaussian Copula recovered as a special case when
$\nu \rightarrow \infty$. With three inputs to replace, we can still
capture the entire dependence structure with a $3 \times 3$
correlation matrix whose estimates are defensible by physical science
literature, now augmented by a single additional parameter $\nu$ that
we explore through sensitivity analysis.

### Shape of distribution for each hazard

**Tornado Annualized Frequency:** We chose the Poisson distribution
for this hazard because the occurrence are discrete counts of whole
number and last year's chance of experiencing a tornado does not
change the next years. The Poisson Distribution is also the natural
model for data like this because it is designed to model the number of
times a random, rare event occurs within a specific interval of time.
The Poisson Distribution is governed by a single parameter $\lambda$,
which is the average number of events per year. There is only one
parameter to estimate, making this straightforward even with sparse
historical data. We will also check whether the zero-inflated
variation fits the historical data better because of the excess amount
of 0s.

**Flood Historical Loss Ratio:** We used a two-part hurdle model
because the flood loss in Manhattan is 0 in most years, then
occasionally very large. Given the nature of the variable, we wish to
seperate the if, and the how much, so there's two parts to the
distribution 1. Did a flood happen this year? 2. Given that a loss was
recorded, how severe is it? We fit a lognormal distribution to these
positive loss years.

**Earthquake Historical Loss Ratio:** Similarly, we used two-part
structure as the above. But because in our period of record, there was
no recorded earthquakes, so we cannot fit a distribution to zero
observations. Instead, we set the parameters based on published
structural engineering studies: the USGS seismic hazard maps, FEMA's
HAZUS earthquake loss tables, and Tantala et al. (2003), who estimated
earthquake losses for the NYC metro area.

## Part 2: Estimating parameters for Tornado Annual Frequency

### Data Source and Marginal Model

Annual tornado event counts for New York County (FIPS 36061) are drawn
from SHELDUS v19 (Hazard = "Tornado"), which itself sources from the
NOAA Storm Events Database. The period of record is 1960–2022 (63
years): <https://www.ncdc.noaa.gov/stormevents/>
<https://cemhs.asu.edu/sheldus>

**Marginal Model** Poisson distribution, if $\lambda_T = 0.1$, that
means on average one tornado per decade. The distribution tells us
exactly how likely 0, 1, 2, 3, ... tornadoes are in any given year,
all from that single parameter. Formally:
$N_T \sim \text{Poisson}(\lambda_T)$.

### Annual Count

SHELDUS records only event years, so years with zero tornadoes are
absent from the dataset. We can reconstruct the full annual count
series by joining onto a complete year and filling absences with zero.

```{r}
#| label: tornado-series

por_start_trnd <- 1960
por_end_trnd   <- 2022
por_years_trnd <- por_end_trnd - por_start_trnd + 1

tornado_events <- sheldus_county_clean %>%
  filter(Hazard == "Tornado") %>%
  group_by(Year) %>%
  summarise(n_events = n(), .groups = "drop")

annual_tornado <- tibble(Year = por_start_trnd:por_end_trnd) %>%
  left_join(tornado_events, by = "Year") %>%
  mutate(n_events = replace_na(n_events, 0L))

tibble(
  Statistic = c("Period of record", "Total years", "Observed events",
                "Years with ≥ 1 event", "Years with zero events"),
  Value     = c(paste(por_start_trnd, "–", por_end_trnd),
                por_years_trnd,
                sum(annual_tornado$n_events),
                sum(annual_tornado$n_events > 0),
                sum(annual_tornado$n_events == 0))
) %>%
  kable(caption = "Tornado annual count series — summary statistics") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

### Exploratory Analysis - Tornado Data

```{r}
#| label: tornado-explore

# When did tornadoes actually occur?
annual_tornado %>%
  filter(n_events > 0) %>%
  print()

# Frequency table of annual counts
print(table(annual_tornado$n_events))

# Timeline plot
ggplot(annual_tornado, aes(x = Year, y = n_events)) +
  geom_col(fill = "steelblue", width = 0.8) +
  labs(
    title    = "Annual Tornado Count — Manhattan 1960–2022",
    subtitle = "Only 3 tornado events recorded across 63 years",
    x = "Year", y = "Number of tornado events"
  ) +
  theme_minimal(base_size = 12)

# Sample mean and sample variance 
cat("Mean:             ", round(mean(annual_tornado$n_events), 4), "\n")
cat("Variance:         ", round(var(annual_tornado$n_events),  4), "\n")
cat("Ratio (var/mean): ", round(var(annual_tornado$n_events) /
                                  mean(annual_tornado$n_events), 4), "\n")
```

### MLE for Poisson ($\lambda_T$)

Through the Maximum Likelihood Estimation (MLE) method, we want to
answer the question of, what value of the parameter makes the data
actually observed as probable as possible? For the Poisson
distribution, we seek to maximize the log-likelihood function: $$
\ell(\lambda) = \sum_{t=1}^{T} \left[ x_t \log\lambda - \lambda - \log(x_t!) \right]
$$

**Source:** Taboga, Marco (2021). "Poisson distribution - Maximum
Likelihood Estimation", Lectures on probability theory and
mathematical statistics. Kindle Direct Publishing. Online appendix.
https://www.statlect.com/fundamentals-of-statistics/Poisson-distribution-maximum-likelihood.

Setting its derivative to 0 gives us $\hat{\lambda} = \bar{x}$, which
means the estimator $\hat{\lambda}$ is just the sample mean (proof
given in the above source). And the standard error tells us how far,
on average, a sample mean is likely to differ from the true population
mean. $$
\hat{\lambda} = \bar{x}, \qquad \text{SE}(\hat{\lambda}) = \sqrt{\hat{\lambda}/T}
$$

```{r}
#| label: tornado-mle

# Estimating parameter lambda hat
lambda_hat <- mean(annual_tornado$n_events)

# Standard error of lambda hat
lambda_se  <- sqrt(lambda_hat / por_years_trnd)

# Confidence Interval - because for large samples like ours, the MLE lambda hat is asymptotically normally distributed (proof in source above). So +- 1.96 is used to calculate the 95% confidence interval
lambda_ci  <- lambda_hat + c(-1.96, 1.96) * lambda_se

tibble(
  Parameter        = c("λ̂ (annual rate)", "SE(λ̂)", "95% CI lower",
                        "95% CI upper", "Mean return period (years)"),
  Value            = round(c(lambda_hat, lambda_se, lambda_ci[1],
                              lambda_ci[2], 1 / lambda_hat), 4)
) %>%
  kable(caption = "Poisson MLE — tornado annual frequency, Manhattan 1960–2022") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

# Save estimates for for simulation process
lambda_T <- lambda_hat 
```

### Goodness-of-fit

With the parameters estimated, we compare our data observed tornado
annual frequencies against the Poisson-predicted frequencies. With
only 3 observed tornado years in our data, a simple visual inspection
is used.

**Source:
https://www.nku.edu/~longa/Rweb/library/base/html/Poisson.html#:~:text=dpois%20gives%20the%20density%2C%20ppois,and%20rpois%20generates%20random%20deviates.**

```{r}
#| label: tornado-gof

# find the highest number of tornadoes recorded in a single year within your dataset
max_count <- max(annual_tornado$n_events)

# counts how many years had 0 tornadoes, 1 tornado, etc but only counts positive integers (+1L)
obs_freq  <- tabulate(annual_tornado$n_events + 1L, nbins = max_count + 1)

# calculates the probability of seeing each tornado count (0, 1, 2...) based on your average rate (lambda hat, assuming a perfect poisson distribution
exp_freq  <- dpois(0:max_count, lambda_hat) * por_years_trnd

tibble(
  `Annual count`    = 0:max_count,
  `Observed years`  = obs_freq,
  `Expected (Poisson)` = round(exp_freq, 2)
) %>%
  kable(caption = "Observed vs. Poisson-expected annual tornado frequencies — Manhattan") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

### Zero-inflated Poisson Comparison

Since our sample has 60 non-tornado years and 3 tornado years, we
check to see if the zero-inflated Poisson is a better fit because it
adds another explanation for zeros (maybe some years are structurally
tornado-free regardless of the real chance of tornado occurrence
instead of a non-tornado year where it just happened that no tornado
occurred).

We compare the two models using Akaike Information Criterion (AIC),
which estimates the quality of each model relative to the other, given
a set of data. Let $k$ be the number of parameters needed and
$\hat{L}$ be the maximized value of the model's likelihood function,
the AIC value of the model is

$$
AIC = 2k-2ln(\hat{L}).
$$

**Source:** Wikipedia contributors, "Akaike information criterion,"
Wikipedia, The Free Encyclopedia,
https://en.wikipedia.org/w/index.php?title=Akaike_information_criterion&oldid=1319366559
(accessed April 11, 2026).

This measure rewards a model for its goodness of fit but penalizes it
for amount of parameter it needs. We choose the model with a lower
AIC, so if the Zero Inflated Poisson model does not meaningfully
improve goodness-of-fit despite needing an extra parameter, we will
use the simpler Poisson model.

```{r}
#| label: tornado-zip-comparison

# two-part formula separated by the pipe (|) where n_events ~ 1 is the count part (models the number of events across all years in dataset rather than predicting based on other variables). And "| 1" is the Zero-inflation part (Models the probability of being an "excess zero" using only an intercept).
zip_fit <- pscl::zeroinfl(n_events ~ 1 | 1, data = annual_tornado, dist = "poisson")

# calculate the overall average rate of n_events across all years in the annual_tornado dataset, rather than predicting it based on other variables (arg 1)
poi_fit <- glm(n_events ~ 1, data = annual_tornado, family = poisson)

tibble(
  Model  = c("Poisson", "Zero-Inflated Poisson"),
  AIC    = round(c(AIC(poi_fit), AIC(zip_fit)), 2),
  `# Parameters` = c(1L, 2L)
) %>%
  kable(caption = "Poisson vs. ZIP model comparison — lower AIC is preferred") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

We will proceed with the Poisson because of its lower AIC value.

### Overdispersion

Because a Poisson Distribution assumes variance is equal to the mean,
which allows the distribution to be defined using only 1 parameter. We
want to test if this is too bold of an assumption, so we test for
overdispersion. Based on this result, we will not switch to fitting
the tornado count data to a Negative Binomial distribution.

**Source:**

1.  https://www.rdocumentation.org/packages/performance/versions/0.12.4/topics/check_overdispersion
2.  https://rpubs.com/DonArres/TestingOverdispersion

```{r}
#| label: tornado-overdispersion-raw
# Overdisperson in our sample, which we got 0.9677
raw_mean <- mean(annual_tornado$n_events)
raw_var  <- var(annual_tornado$n_events)
raw_ratio <- raw_var / raw_mean

tibble(
  Statistic = c("Sample Mean", "Sample Variance", "Dispersion Ratio (Var/Mean)"),
  Value     = round(c(raw_mean, raw_var, raw_ratio), 4)
) %>%
  kable(caption = "Raw overdispersion check — annual tornado counts before model fitting") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r}
# testing overdispersion for fitted Poisson, resulted in 0.968, which is pretty close to our ratio from historical data
check_overdispersion(poi_fit)
```

## Part 3: Estimating Parameters for Annual Flood Loss

```{r}
#| label: flood-records-manhattan

# Check raw records before any pre-processing
cat("Raw SHELDUS flood records (Manhattan): \n")
sheldus_raw %>%
  filter(County_FIPS == fips_county) %>%
  filter(Hazard %in% c("Flooding", "Coastal")) %>%
  select(Year, Month, Hazard, prop_dmg_adj, crop_dmg_adj,
         Fatalities, Injuries, Duration_Days) %>%
  print(n = 50)

# Check cleaned/aggregated records
cat("Cleaned SHELDUS flood records (Manhattan): \n")
sheldus_county_clean %>%
  filter(Hazard %in% c("Flooding", "Coastal")) %>%
  mutate(has_loss = prop_dmg_adj > 0) %>%
  select(Year, Month, Hazard, prop_dmg_adj, has_loss) %>%
  print(n = 50)

# Summary breakdown
cat("Breakdown: occurrence vs. damage: \n")
sheldus_county_clean %>%
  filter(Hazard %in% c("Flooding", "Coastal")) %>%
  mutate(has_loss = prop_dmg_adj > 0) %>%
  count(has_loss) %>%
  print()

# Annual aggregation
cat("Annual flood records: \n")
sheldus_county_clean %>%
  filter(Hazard %in% c("Flooding", "Coastal")) %>%
  group_by(Year) %>%
  summarise(n_records  = n(),
            total_loss = sum(prop_dmg_adj, na.rm = TRUE),
            .groups    = "drop") %>%
  arrange(desc(n_records)) %>%
  print(n = 50)

flood_events <- sheldus_county_clean %>%
  filter(Hazard %in% c("Flooding", "Coastal")) %>%
  group_by(Year) %>%
  summarise(n_records  = n(),
            total_loss = sum(prop_dmg_adj, na.rm = TRUE),
            .groups    = "drop")

annual_flood_spine <- tibble(Year = por_start_trnd:por_end_trnd) %>%
  left_join(flood_events, by = "Year") %>%
  mutate(n_records  = replace_na(n_records,  0L),
         total_loss = replace_na(total_loss, 0))

cat("Total years in period:              ", por_years_trnd, "\n")
cat("Years with any flood record:        ",
    sum(annual_flood_spine$n_records > 0), "\n")
cat("Years with flood damage > $0:       ",
    sum(annual_flood_spine$total_loss > 0), "\n")
cat("Years with zero records:            ",
    sum(annual_flood_spine$n_records == 0), "\n")

# Show only the years where something happened
cat("Years with at least one flood record: \n")
annual_flood_spine %>%
  filter(n_records > 0) %>%
  mutate(total_loss = dollar(round(total_loss))) %>%
  print(n = 50)
```

### Data Source and Marginal Model

Annual flood losses for New York County (FIPS 36061) are drawn from
SHELDUS v19. <https://cemhs.asu.edu/sheldus>

**Marginal Model** Two-part hurdle model with lognormal distribution.
Loss data is fundamentally different from a loss data because losses
are only to be incurred if the hazard was present to begin with. So a
two part hurdle model combines a binary model to model whether or not
flooding happened and a lognormal model to see how severe the damage
was, given there was a flooding event.

Formally, let $L_F$ be the annual loss from flood events, then our
data is seperated into years with no loss means $L_F=0$ and years with
loss $L_F>0$. Consider when $L_F=0$, let $p_f=P(L_F>0)$ be the
probability that any flood damages occur in a given year. Consider
when $L_F>0$, we fit their distribution to a lognormal distribution
with parameters $\mu_F$ (mean of underlying distribution) and
$\sigma_F$ (standard deviation of distribution). We chose a lognormal
distribution because flood damages tend to grow multiplicatively where
a moderate flood might cause 10x the damage as a mild one, and a major
flood might cause 10x the damage of a moderate one and so on. Thus, it
is likely that the log transformed loss of flood damage is normally
distributed. Combining the two gives us

$$
L_F = \begin{cases} 0 & \text{with probability } 1 - p_F \\ \text{Lognormal}~(\mu_F, \sigma^2_F) & \text{with probability } p_F \end{cases}
$$

**Source:**

1.  Cragg, J. G., "Some Statistical Models for Limited Dependent
    Variables with Application to the Demand for Durable Goods,"
    Econometrica, vol. 39, no. 5, pp. 829–844, 1971.
    https://doi.org/10.2307/1909582

2.  Wooldridge, J. M., "Hurdle and Selection Models," BGSE/IZA Course
    in Microeconometrics Lecture Notes, Michigan State
    University, 2009.
    https://legacy.iza.org/teaching/wooldridge-course-09/course_html/docs/slides_twopart_5_r1.pdf
    (accessed April 11, 2026).

3.  Hsu, A.-C. and Liu, S.-C., "The hurdle models choice between
    truncated normal and lognormal," Applied Economics, vol. 40, no.
    2, pp. 201–207, 2008. https://doi.org/10.1080/00036840600749581

### Annual Loss

```{r}
#| label: flood-annual-loss

annual_flood <- sheldus_county_clean %>%
  filter(Hazard %in% c("Flooding", "Coastal")) %>%
  group_by(Year) %>%
  summarise(total_loss = sum(prop_dmg_adj, na.rm = TRUE), .groups = "drop")

annual_flood_full <- tibble(Year = por_start_trnd:por_end_trnd) %>%
  left_join(annual_flood, by = "Year") %>%
  mutate(total_loss = replace_na(total_loss, 0))

flood_loss_years   <- annual_flood_full %>% filter(total_loss > 0)
n_flood_loss_years <- nrow(flood_loss_years)

tibble(
  Statistic = c("Period of record", "Years with positive loss",
                "Years with zero loss", "Maximum single-year loss"),
  Value     = c(paste(por_start_trnd, "–", por_end_trnd),
                n_flood_loss_years,
                por_years_trnd - n_flood_loss_years,
                dollar(round(max(annual_flood_full$total_loss))))
) %>%
  kable(caption = "Annual flood loss series — summary statistics, Manhattan") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

### Explanatory Analysis - Flood Data

```{r}
#| label: flood-explore
#| fig-height: 5

# Timeline of annual flood losses
ggplot(annual_flood_full, aes(x = Year, y = total_loss / 1e6)) +
  geom_col(fill = "steelblue", width = 0.8) +
  labs(
    title    = "Annual Flood Building Loss — Manhattan 1960–2022",
    subtitle = "Zero in most years; sporadic moderate losses when events occur",
    x = "Year", y = "Loss ($M, 2022 adj.)"
  ) +
  theme_minimal(base_size = 12)

# Raw vs. log-transformed positive losses side by side
if (n_flood_loss_years >= 3) {
  p_raw <- ggplot(flood_loss_years, aes(x = total_loss / 1e6)) +
    geom_histogram(bins = 12, fill = "steelblue",
                   color = "white", alpha = 0.8) +
    labs(title = "Raw losses ($M)", x = "Loss ($M)", y = "Count") +
    theme_minimal(base_size = 11)

  p_log <- ggplot(flood_loss_years, aes(x = log(total_loss))) +
    geom_histogram(bins = 12, fill = "coral",
                   color = "white", alpha = 0.8) +
    labs(title = "Log(losses)",
         x = "log(Loss)", y = "Count") +
    theme_minimal(base_size = 11)

  print(p_raw + p_log +
    plot_annotation(
      title    = "Flood severity: raw vs. log-transformed positive-loss years",
      subtitle = paste0(n_flood_loss_years, " positive-loss years out of ",
                        por_years_trnd, " total")
    ))

  # Summary of positive-loss years
  cat("Positive-loss years: summary")
  print(summary(flood_loss_years$total_loss))
  cat("Empirical p_F:", round(p_F_hat <- n_flood_loss_years / por_years_trnd, 3), "\n")
}
```

### Part (a) Bernoulli MLE

To estimate $p_F$, notice the maximum likelihood estimate for a
Bernoulli distribution is the sample mean, given $T$ independent draws
$$
\hat{p}_F = \frac{\text{# years with loss} > 0}{T};
$$ \$\$

\text{SE}(\hat{p}\_F) = \sqrt{\frac{\hat{p}_F(1-\hat{p}_F)}{T}}. \$\$

**Sources:**

1.  Casella, G. and Berger, R. L., Statistical Inference, 2nd ed.,
    Duxbury Press, Pacific Grove, CA, 2002, pp. 315–320 (MLE), pp.
    347–349.
2.  Wasserman, L., All of Statistics: A Concise Course in Statistical
    Inference, Springer, New York, 2004, pp. 119–121.
    https://doi.org/10.1007/978-0-387-21736-9
3.  Wikipedia contributors, "Bernoulli distribution," Wikipedia, The
    Free Encyclopedia,
    https://en.wikipedia.org/wiki/Bernoulli_distribution (accessed
    April 11, 2026).

```{r}
#| label: flood-bernoulli

p_F_hat <- n_flood_loss_years / por_years_trnd
p_F_se  <- sqrt(p_F_hat * (1 - p_F_hat) / por_years_trnd)

tibble(
  Parameter = c("p̂_F (prob. positive-loss year)", "SE(p̂_F)",
                 "95% CI lower", "95% CI upper"),
  Value     = round(c(p_F_hat, p_F_se,
                       p_F_hat - 1.96 * p_F_se,
                       p_F_hat + 1.96 * p_F_se), 4)
) %>%
  kable(caption = "Bernoulli MLE — flood occurrence probability, Manhattan") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

### Part (b) Lognormal MLE for Severity

For years where flood loss occurred, we fit the loss severity to a
lognormal distribution by estimating two parameters, $\mu_F$ (the mean
of logged losses) and $\sigma_F$ (the spread of logged losses).

```{r}
#| label: flood-lognormal

  flood_fit <- fitdistrplus::fitdist(flood_loss_years$total_loss,
                                     distr = "lnorm", method = "mle")
  mu_F    <- flood_fit$estimate["meanlog"]
  sigma_F <- flood_fit$estimate["sdlog"]

tibble(
  Parameter = c("μ̂_F (meanlog)", "σ̂_F (sdlog)",
                 "Implied median ($)", "Implied mean ($)",
                 "Implied 95th percentile ($)", "Implied 99th percentile ($)"),
  Value     = c(round(mu_F, 4),
                round(sigma_F, 4),
                dollar(round(exp(mu_F))),
                dollar(round(exp(mu_F + 0.5 * sigma_F^2))),
                dollar(round(qlnorm(0.95, mu_F, sigma_F))),
                dollar(round(qlnorm(0.99, mu_F, sigma_F))))
) %>%
  kable(caption = "Lognormal MLE — flood severity conditional on loss > 0 for Manhattan") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

### Goodness-of-fit

We check how well a lognormal distribution fits our loss data through
the Anderson-Darling test for normality. We chose this test over
various other normality tests because tail behavior is the main focus
of our work, and A-D test places more weight on tail departures,
making it the best for detecting outliers and heavy-tails.

The null hypothesis being tested is:
$H_0: \log(L_F) \sim N(\mu, \sigma^2)$.

```{r}
#| label: flood-ad-test
  log_losses <- log(flood_loss_years$total_loss)
  ad_result  <- nortest::ad.test(log_losses)

  tibble(
    Test      = "Anderson-Darling",
    Statistic = round(ad_result$statistic, 4),
    `p-value` = round(ad_result$p.value,   4),
    Conclusion = ifelse(ad_result$p.value > 0.05,
                        "Fail to reject H₀ — Lognormal adequate",
                        "Reject H₀ — consider GPD alternative")
  ) %>%
    kable(caption = "Anderson-Darling GOF test — log(annual flood loss), Manhattan") %>%
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

We also plot a Quantile-Quantile (QQ) plot to assess if our data
follows our proposed lognormal distribution by plotting the quantiles
of sample data against lognormal quantiles. The points fall along a
straight 45-degree line, meaning the distributions are in fact
similar.

```{r}
  #|label: Q-Q plot
  qqnorm(log_losses,
         main = "Q-Q Plot: log(Annual Flood Loss) — Manhattan",
         xlab = "Theoretical Normal Quantiles (z-scores)",
         ylab = "log(Annual Flood Loss in USD)",
         pch  = 19, col = "steelblue", cex = 0.85)
  qqline(log_losses, col = "firebrick", lwd = 1.5)
```

### Complete Fitted Distribution

The first graph shows the Bernoulli split (what fraction of years have
any loss). The second shows the Lognormal density for the severity
conditional on a loss occurring.

```{r}
#| label: flood-dist-plots
#| fig-height: 8

# Part A: Bernoulli bar chart
p_bern <- tibble(
  outcome = c("No loss (0)", "Loss occurs"),
  prob    = c(1 - p_F_hat, p_F_hat)
) %>%
  ggplot(aes(x = outcome, y = prob, fill = outcome)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = percent(prob, accuracy = 0.1)),
            vjust = -0.4, size = 4) +
  scale_fill_manual(values = c("No loss (0)"  = "grey70",
                               "Loss occurs"  = "steelblue")) +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
  labs(
    title    = "Flood Part 1 — Bernoulli occurrence model",
    subtitle = paste0("p̂_F = ", round(p_F_hat, 3),
                      ": ~", round(p_F_hat*100), "% of years have any flood loss"),
    x = NULL, y = "Probability"
  ) +
  theme_minimal(base_size = 12)

# Part B: Lognormal severity curve
x_fl   <- exp(seq(log(1e4), log(2e9), length.out = 600))
dens_fl <- dlnorm(x_fl, meanlog = mu_F, sdlog = sigma_F)

p_ln <- ggplot(data.frame(x = x_fl / 1e6, y = dens_fl), aes(x, y)) +
  geom_line(color = "steelblue", linewidth = 1.1) +
  geom_vline(xintercept = exp(mu_F) / 1e6,
             linetype = "dashed", color = "firebrick", linewidth = 0.9) +
  geom_vline(xintercept = exp(mu_F + 0.5 * sigma_F^2) / 1e6,
             linetype = "dotted", color = "darkgreen", linewidth = 0.9) +
  annotate("text",
           x = exp(mu_F) / 1e6 * 1.15,
           y = max(dens_fl) * 0.75,
           label = paste0("Median\n$", round(exp(mu_F) / 1e3), "K"),
           color = "firebrick", hjust = 0, size = 3.2) +
  annotate("text",
           x = exp(mu_F + 0.5 * sigma_F^2) / 1e6 * 1.15,
           y = max(dens_fl) * 0.45,
           label = paste0("Mean\n$", round(exp(mu_F + 0.5*sigma_F^2) / 1e6, 1), "M"),
           color = "darkgreen", hjust = 0, size = 3.2) +
  scale_x_log10(labels = dollar_format(suffix = "M")) +
  labs(
    title    = "Flood Part 2 — Lognormal severity (conditional on loss > 0)",
    subtitle = paste0("μ̂ = ", round(mu_F, 2), ", σ̂ = ", round(sigma_F, 2),
                      " | mean far above median = fat right tail"),
    x = "Annual flood loss (log scale, $M)", y = "Density"
  ) +
  theme_minimal(base_size = 12)

print(p_bern / p_ln)
```

## Part 4: Estimating Parameters for Earthquake Building Loss

### Data Source and Marginal Model

```{r}
#|labels: Earthquake-records-manhattan 

# Check raw records before any pre-processing
cat("Raw SHELDUS earthquake records (Manhattan): \n")
sheldus_raw %>%
  filter(County_FIPS == fips_county) %>%
  filter(grepl("earthquake|quake", Hazard, ignore.case = TRUE)) %>%
  select(Year, Month, Hazard, prop_dmg_adj, crop_dmg_adj,
         Fatalities, Injuries, Duration_Days) %>%
  print(n = 50)

# Check cleaned/aggregated records
cat("Cleaned SHELDUS earthquake records (Manhattan): \n")
sheldus_county_clean %>%
  filter(grepl("earthquake|quake", Hazard, ignore.case = TRUE)) %>%
  mutate(has_loss = prop_dmg_adj > 0) %>%
  select(Year, Month, Hazard, prop_dmg_adj, has_loss) %>%
  print(n = 50)

# Summary breakdown 
cat("Breakdown: occurrence vs. damage: \n")
sheldus_county_clean %>%
  filter(grepl("earthquake|quake", Hazard, ignore.case = TRUE)) %>%
  mutate(has_loss = prop_dmg_adj > 0) %>%
  count(has_loss) %>%
  print()

# Annual aggregation 
cat("Annual earthquake records: \n")
sheldus_county_clean %>%
  filter(grepl("earthquake|quake", Hazard, ignore.case = TRUE)) %>%
  group_by(Year) %>%
  summarise(n_records  = n(),
            total_loss = sum(prop_dmg_adj, na.rm = TRUE),
            .groups    = "drop") %>%
  arrange(desc(n_records)) %>%
  print(n = 50)

eq_events <- sheldus_county_clean %>%
  filter(grepl("earthquake|quake", Hazard, ignore.case = TRUE)) %>%
  group_by(Year) %>%
  summarise(n_records  = n(),
            total_loss = sum(prop_dmg_adj, na.rm = TRUE),
            .groups    = "drop")

annual_eq_full <- tibble(Year = por_start_trnd:por_end_trnd) %>%
  left_join(eq_events, by = "Year") %>%
  mutate(n_records  = replace_na(n_records,  0L),
         total_loss = replace_na(total_loss, 0))

cat("Total years in period:", por_years_trnd, "\n")
cat("Years with any earthquake record:   ",
    sum(annual_eq_full$n_records > 0), "\n")
cat("Years with earthquake damage > $0:  ",
    sum(annual_eq_full$total_loss > 0), "\n")
cat("Years with zero records:            ",
    sum(annual_eq_full$n_records == 0), "\n")


# Show only the years where something happened
cat("Years with at least one earthquake record: \n")
annual_eq_full %>%
  filter(n_records > 0) %>%
  mutate(total_loss = dollar(round(total_loss))) %>%
  print(n = 50)
```

Manhattan has not experienced earthquake damage in our 63 years of
record, so we will not be able to estimate occurrence (and hence loss)
probability from no observed data. So we use the standard solution of
using external evidence like published engineering studies to set
parameters for our distribution.

**Marginal model:** For the same reasons outlined above, we will use
the same two-part hurdle structure as annual flood loss because we ask
the same two questions, did a damaging earthquake occur this year? If
so, how much was lost? We use published resources to get the
appropriate parameter rather than fitting the distribution to our
SHELDUS data:

1.  **USGS National Seismic Hazard Maps (2023).** At the 2%
    probability of exceedance in 50 years (2,475-year return period),
    the tool returns a PGA of 0.1803g (in any given year, there's a 1
    in 2475 chance of exceeding this shake level). This is about 18%
    of the acceleration of gravity, which is strong but generally not
    destructive to well-engineered modern structures. We use this to
    anchor the tail behavior of $L_E$.(USGS Unified Hazard Tool,
    Conterminous U.S. 2014 (v4.0.x), Site Class B/D, Manhattan:
    40.7831N, 73.9712W)
    <https://earthquake.usgs.gov/hazards/interactive/>

2.  **NYC Hazard Mitigation Plan (NYC OEM, 2024)**: said there was a
    "5 to 25 percent chance of a slight or damaging earthquake
    occurring in the New York City area over a 100-year period." Using
    the midpoint (15% in 100 years) to convert to an annual
    probability gives us 1 - (1 - 0.15)\^(1/100) ≈ 0.0016/year.
    <https://nychazardmitigation.com/documentation/hazard-profiles/earthquakes/>

3.  **Tantala, M.W., Gustafson, J.L., Deodatis, G., & Jacob, K.
    (2008).** "Earthquake loss estimation for the New York City
    Metropolitan Region." *Soil Dynamics and Earthquake Engineering*,
    28(9), 812–835. Under the 2,500-year return period probabilistic
    scenario (2% probability of exceedance in 50 years), Tantala
    estimate New York County direct building loss at *4.21B* (Fig.
    12F). Note that the M5 scenario produces a maximum
    per-census-tract loss of only \$58,955 (Fig. 12A), which confirms
    that moderate-magnitude events near Manhattan generate minimal
    building damage and that tail risk is likely concentrated in rare
    but high-intensity events.

### Estimate parameter values

We estimate parameter values using these literatures.

```{r}
#| label: earthquake-params
# Event probability - NYC Hazrd Mitigation plan estiated there was a 5-25% chance of slight or damaging earthquake occuring in the NYC area over 100 years. Although not directly about Manhattan, an earthquake that hits NYC in general will surely effect Manhattan. We use the midpoint of the estimate to arrive at an estimated annual earthquake occurrence probability of 0.0016/year. With slight round up to account for the difficulty in estimating seismic events.In sensitivity analysis, we will account for the full 5-25% range to check for robustness.
p_E <- 0.002

# Conditional severity - LogNormal(mu_E, sigma_E^2)
# We choose the parameter so that the 99.9th percentile of the conditional lognormal approximates New York County direct building loss under the 2,500-year probabilistic scenario. We choose sigma_E to reflect the large spread between small-event losses and severe scenario losses in the literature. Based on this choice, we then set mu_E such that the 99.9th percentile of the conditional severity distribution is approximately aligned with the $4.21B New York County scenario loss from Tantala et al. (2008).
# Notice this is a choice statistical estimate from observed data. Also, for our constraint of 4.21B loss to be met, there are infinitely many (sigma_E, mu_E) that satisfies. Within reason, we should set sigma_E between 0.9 and 1.5 because it controls how widely earthquake losses can vary around the "typical"average" outcome. A value near 0.9 will mean relatively limited dispersion in losses, while a value near 1.5 allows for a much heavier tail, which is useful for capturing rare but very large building-loss yearsWe choose the midpoint 1.2, and we test 0.9 and 1.5 in the sensitivity analysis.
sigma_E <- 1.2
mu_E <- log(4.21e9) - qnorm(0.999) * sigma_E  

# verification table
tibble(
  Parameter = c(
    "p_E (annual prob. of damaging event)",
    "μ_E (meanlog of conditional severity)",
    "σ_E (sdlog of conditional severity)",
    "Implied median loss ($)",
    "Implied 95th percentile ($)",
    "Implied 99th percentile ($)",
    "Implied 99.9th percentile ($)"
  ),
  Value = c(
    p_E,
    mu_E,
    sigma_E,
    dollar(round(exp(mu_E))),
    dollar(round(qlnorm(0.95, mu_E, sigma_E))),
    dollar(round(qlnorm(0.99, mu_E, sigma_E))),
    dollar(round(qlnorm(0.999, mu_E, sigma_E)))
  ),
  Source = c(
    "NYC Hazard Mitigation Plan (2024)",
    "Calibrated to Tantala et al. (2008), Fig. 12F",
    "Chosen to reflect heavy-tailed building-loss variation across scenarios",
    "—", "—", "—", "—"
  )
) %>%
  kable(caption = "Earthquake parameters — literature-informed priors for Manhattan") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

### Distribution Plot

Since there are no observed losses to explore and compare our fitted
distribution to, we show what the literature-derived severity
distribution for earthquake building loss looks like to check if the
parameters produce reasonable implied losses.

```{r}
#| label: earthquake-explore
#| fig-height: 10

# Part A: Bernoulli — how rare is a damaging event?
p3a <- tibble(
  outcome = c("No damage (0)", "Damaging event"),
  prob    = c(1 - p_E, p_E)
) %>%
  ggplot(aes(x = outcome, y = prob, fill = outcome)) +
  geom_col(width = 0.5, show.legend = FALSE) +
  geom_text(aes(label = percent(prob, accuracy = 0.001)),
            vjust = -0.4, size = 4) +
  scale_fill_manual(values = c("No damage (0)"  = "grey70",
                               "Damaging event" = "firebrick")) +
  scale_y_continuous(labels = percent_format(), limits = c(0, 1)) +
  labs(
    title    = "Earthquake Part 1 — Bernoulli occurrence",
    subtitle = paste0("p_E = ", p_E,
                      ": damaging event roughly once every 500 years"),
    x = NULL, y = "Probability"
  ) +
  theme_minimal(base_size = 12)

# Part B: Lognormal severity, if there is a earthquake
x_eq   <- exp(seq(log(1e8), log(5e10), length.out = 600))
dens_eq <- dlnorm(x_eq, meanlog = mu_E, sdlog = sigma_E)

p3b <- ggplot(data.frame(x = x_eq / 1e9, y = dens_eq), aes(x, y)) +
  geom_line(color = "firebrick", linewidth = 1.1) +
  geom_vline(xintercept = exp(mu_E) / 1e9,
             linetype = "dashed", color = "firebrick", linewidth = 0.9) +
  annotate("text",
           x = exp(mu_E) / 1e9 * 1.08,
           y = max(dens_eq) * 0.8,
           label = paste0("Median\n$", round(exp(mu_E) / 1e9, 1), "B"),
           color = "firebrick", hjust = 0, size = 3.5) +
  geom_vline(xintercept = exp(mu_E + 0.5 * sigma_E^2) / 1e9,
           linetype = "dotted", color = "darkgreen", linewidth = 0.9) +
  annotate("text",
         x = exp(mu_E + 0.5 * sigma_E^2) / 1e9 * 1.08,
         y = max(dens_eq) * 0.5,
         label = paste0("Mean\n$", round(exp(mu_E + 0.5 * sigma_E^2) / 1e9, 1), "B"),
         color = "darkgreen", hjust = 0, size = 3.5) +
  scale_x_log10(labels = dollar_format(suffix = "B")) +
  labs(
    title    = "Earthquake Part 2 — Lognormal severity (conditional on event)",
    subtitle = paste0("μ_E = ", round(mu_E,3), ", σ_E = ", sigma_E),
    x = "Conditional building loss ($B, log scale)", y = "Density"
  ) +
  theme_minimal(base_size = 12)

# Percentile table for quick sanity check
cat("Earthquake severity percentile table")
tibble(
  Percentile = c("50th (median)", "90th", "95th", "99th", "99.9th"),
  `Conditional loss` = dollar(round(
    qlnorm(c(0.50, 0.90, 0.95, 0.99, 0.999), mu_E, sigma_E)
  ))
) %>% print()

print(p3a / p3b)
```

## Part 5: Specify t-Copula

We pick the t-Copula because it outperforms the Gaussian in modeling
extreme joint risks due to its heavier tail dependence. The main input
we need for the t-Copula is the $3\times 3$ matrix, $\mathbf{P}$ of
pairwise correlation between the occurrence of three hazards.

|            | Tornado | Flood | Earthquake |
|------------|---------|-------|------------|
| Tornado    | 1.00    | 0.05  | 0.00       |
| Flood      | 0.05    | 1.00  | 0.00       |
| Earthquake | 0.00    | 0.00  | 1.00       |

We also need the degree of freedom as the other input, in this case,
we picked 4 due to our hypothesized heavy tail dependence.
**Justification:**

1.  $\rho(\text{Tornado}, \text{Flood}) = 0.05$ (weak positive
    correlation): The empirical Spearman correlation (assess when
    tornado count is higher, does flood loss tend to be higher too?)
    between annual tornado count and annual flood building kiss un
    Manhattan in our period of record is about 0.035. We choose to set
    $\rho_{TF}=0.05$, which overstates the observed dependence from
    our data to acknowledge the theoretical shared convective driver
    of each hazards.

```{r}
corr_check <- tibble(Year = por_start_trnd:por_end_trnd) %>%
  left_join(
    annual_tornado %>% select(Year, n_tornado = n_events),
    by = "Year"
  ) %>%
  left_join(
    annual_flood_full %>% select(Year, flood_loss = total_loss),
    by = "Year"
  ) %>%
  mutate(
    n_tornado  = replace_na(n_tornado,  0),
    flood_loss = replace_na(flood_loss, 0)
  )

# Spearman rank correlation
cat("Spearman correlation (tornado count vs. flood loss):")
cat(round(cor(corr_check$n_tornado, corr_check$flood_loss,
              method = "spearman"), 4), "\n")
```

2.  $\rho(\text{Tornado}, \text{Earthquake}) = 0.00$ and
    $\rho(\text{Flood}, \text{Earthquake}) = 0.00$ because Earthquakes
    are caused by tectonic stress in the earth's crust, which has
    nothing to do with weather. There is no physical reason for a bad
    earthquake year would coincide with a bad tornado or flood year.

### Validate correlation matrix

Recall that for a correlation matrix to be mathematically valid, it
has to be positive definite. We run a quick check to make sure

```{r}
#| label: corr-matrix

rho_TF <- 0.05
rho_TE <- 0.00
rho_FE <- 0.00

P <- matrix(c(1,      rho_TF, rho_TE,
              rho_TF, 1,      rho_FE,
              rho_TE, rho_FE, 1),
            nrow = 3, byrow = TRUE,
            dimnames = list(c("Tornado", "Flood", "Earthquake"),
                            c("Tornado", "Flood", "Earthquake")))

eigenvalues <- eigen(P, symmetric = TRUE)$values
pd_check    <- all(eigenvalues > 0)

if (!pd_check) {
  P <- as.matrix(Matrix::nearPD(P, corr = TRUE)$mat)
  message(" matrix was not positive definite")
}

cat("t-copula correlation matrix P"); print(round(P, 4))
cat("Eigenvalues of P (all must be > 0")
print(round(eigenvalues, 6))
cat("Positive definite:", pd_check)
```

## Part 6: Monte Carlo Simulation

### Setup

We simulate $N=100,000$ hypothetical annual loss years because we want
sufficient samples to ensure we capture the tail behavior of the
simulated risk distribution.

```{r}
#| label: sim-setup

# initialize a pseudorandom number generator: we chose this one because the RNG will give us the same sequence of draws every time. We can choose set.seed(7) or comment it out to get a different (but statistically very similar) set of numbers. We chose 42 for reproducability of the paper and for the graphs and tables to be consistent in the thesis.
set.seed(42)
N_sim <- 100000L
nu <- 4  # degrees of freedom for t-copula

cat("Simulation size:", scales::comma(N_sim), "annual loss years")
```

### Step 1: Draw Copula Sample

We first draw from a tri-variate student's t-distribution with
correlation $\mathbf{P}$ (using the Cholesky factorization of
$\mathbf{P}$), then apply the t-CDF to squish each column into the
$[0,1]$ range. The result $\mathbf{U}$ is our copula sample — 100,000
rows of three correlated uniform values, one column per hazard.

```{r}
#| label: sim-copula-draw




# Using the Cholesky factor L (3x3), we transform T to incorporate the correlation we specified in P (L is used as a transformation).
# W is still a 100,000x3 matrix but columns are now correlated according to our P. Row 1 of W is 1 simulated year of three correlated normal values (1 per hazard)
W <- rmvt(N_sim, sigma = P, df = nu)  

# check the actual sample correlation matrix of W to compare it with our specified P
W_cor <- cor(W)

cat("Sample correlation matrix of W")
print(round(W_cor, 4))
cat("Max absolute deviation from target P:", round(max(abs(W_cor - P)), 5), "\n")


# Now that we have a matrix of correlated normals, we need them to be uniform because the copula is defined on $[0,1]^3$ by the Skylar's theorem. So we apply the normal CDF to mak any real number to a probability distribution from 0 to 1
U <- pt(W, df = nu)
```

As a check, we should see each column of $\mathbf{U}$ should have mean
of around 0.5 and standard deviation of 0.289 (which are mean and SD
of a ndefineduniform \[0,1\]) to check for marginal uniformity

```{r}
tibble(
  Column = c("Tornado (col 1)", "Flood (col 2)", "Earthquake (col 3)"),
  Mean   = round(colMeans(U), 5),
  SD     = round(apply(U, 2, sd), 5),
  `Min`  = round(apply(U, 2, min), 5),
  `Max`  = round(apply(U, 2, max), 5)
) %>%
  kable(caption = "Copula sample U — marginal uniformity check") %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

### Step 2: Convert uniform value into actual loss

Now with $U$ being a uniform random number between 0 and 1,if we feed
it into the inverse CDF of any distribution, we should get random
draws of the distribution. For example, if $U=0.95$, we will get the
95th percentile of the loss distribution we feed it into.

```{r}
# Tornado
sim_tornado <- qpois(U[, 1], lambda = lambda_T)

# Flood
flood_occur <- as.integer(U[, 2] <= p_F_hat)
flood_sev <- qlnorm(runif(N_sim), mu_F, sigma_F) ##Severity drawn independently to make sure when we get sim_eq, it is not limited to the lower end
sim_flood <- flood_occur * flood_sev

# Earthquake
eq_occur <- as.integer(U[, 3] <= p_E)
eq_sev <- qlnorm(runif(N_sim), mu_E, sigma_E)  ##Severity drawn independently to make sure when we get sim_eq, it is not limited to the lower end
sim_eq <- eq_occur * eq_sev

# Summary
bind_rows(
  # Tornado: count variable
  tibble(
    Hazard                    = "Tornado (count)",
    `Pr(event > 0) simulated` = round(mean(sim_tornado > 0), 4),
    `Pr(event > 0) target`    = round(1 - dpois(0, lambda_T), 4),
    `Mean`                    = as.character(round(mean(sim_tornado), 4)),
    `95th pct`                = as.character(round(quantile(sim_tornado, 0.95))),
    `99th pct`                = as.character(round(quantile(sim_tornado, 0.99))),
    `Units`                   = "count"
  ),
  # Flood: conditional on loss occurring
  tibble(
    Hazard                    = "Flood",
    `Pr(event > 0) simulated` = round(mean(sim_flood > 0), 4),
    `Pr(event > 0) target`    = round(p_F_hat, 4),
    `Mean`                    = dollar(round(mean(sim_flood[sim_flood > 0]))),
    `95th pct`                = dollar(round(quantile(sim_flood[sim_flood > 0], 0.95))),
    `99th pct`                = dollar(round(quantile(sim_flood[sim_flood > 0], 0.99))),
    `Units`                   = "$ (conditional on occurrence)"
  ),
  # Earthquake: conditional on loss occurring
  tibble(
    Hazard                    = "Earthquake",
    `Pr(event > 0) simulated` = round(mean(sim_eq > 0), 4),
    `Pr(event > 0) target`    = p_E,
    `Mean`                    = dollar(round(mean(sim_eq[sim_eq > 0]))),
    `95th pct`                = dollar(round(quantile(sim_eq[sim_eq > 0], 0.95))),
    `99th pct`                = dollar(round(quantile(sim_eq[sim_eq > 0], 0.99))),
    `Units`                   = "$ (conditional on occurrence)"
  )
) %>%
  kable(caption = "Simulated marginal distributions — summary statistics") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

**Note:** The above loss statistic for flooding and earthquake is
reported based on their occurrence because they both follow two-part
hurdle model. In first part of their model, we know most years
experience no hazard at all, therefore, no losses. But in the
remaining years, the loss follow a lognormal severity distribution. So
we will get a much more meaningful set of numbers based on the
conditional instead of our summary distribution misleadingly being
flooded with 0 losses. If we don't do this, the summary statistics
will look like the following.

```{r}
#| label: conditional-vs-unconditional

# Unconditional statistics (all 100,000 years including zeros)
uncond <- bind_rows(
  tibble(
    Hazard      = "Flood ($)",
    Basis       = "Unconditional",
    `Pr(> 0)`   = round(mean(sim_flood > 0), 4),
    `Mean`      = dollar(round(mean(sim_flood))),
    `50th pct`  = dollar(round(quantile(sim_flood, 0.50))),
    `95th pct`  = dollar(round(quantile(sim_flood, 0.95))),
    `99th pct`  = dollar(round(quantile(sim_flood, 0.99)))
  ),
  tibble(
    Hazard      = "Earthquake ($)",
    Basis       = "Unconditional",
    `Pr(> 0)`   = round(mean(sim_eq > 0), 4),
    `Mean`      = dollar(round(mean(sim_eq))),
    `50th pct`  = dollar(round(quantile(sim_eq, 0.50))),
    `95th pct`  = dollar(round(quantile(sim_eq, 0.95))),
    `99th pct`  = dollar(round(quantile(sim_eq, 0.99)))
  )
)

# Conditional statistics (positive-loss years only)
cond <- bind_rows(
  tibble(
    Hazard      = "Flood ($)",
    Basis       = "Conditional on occurrence",
    `Pr(> 0)`   = round(mean(sim_flood > 0), 4),
    `Mean`      = dollar(round(mean(sim_flood[sim_flood > 0]))),
    `50th pct`  = dollar(round(quantile(sim_flood[sim_flood > 0], 0.50))),
    `95th pct`  = dollar(round(quantile(sim_flood[sim_flood > 0], 0.95))),
    `99th pct`  = dollar(round(quantile(sim_flood[sim_flood > 0], 0.99)))
  ),
  tibble(
    Hazard      = "Earthquake ($)",
    Basis       = "Conditional on occurrence",
    `Pr(> 0)`   = round(mean(sim_eq > 0), 4),
    `Mean`      = dollar(round(mean(sim_eq[sim_eq > 0]))),
    `50th pct`  = dollar(round(quantile(sim_eq[sim_eq > 0], 0.50))),
    `95th pct`  = dollar(round(quantile(sim_eq[sim_eq > 0], 0.95))),
    `99th pct`  = dollar(round(quantile(sim_eq[sim_eq > 0], 0.99)))
  )
)

# Combined table
bind_rows(uncond, cond) %>%
  arrange(Hazard, Basis) %>%
  kable(
    caption = paste0(
      "Unconditional vs. conditional marginal statistics — ",
      "N = 100,000 simulated years. Unconditional includes all years ",
      "regardless of whether a loss occurred. Conditional restricts to ",
      "years with positive loss only."
    )
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE) %>%
  collapse_rows(columns = 1, valign = "middle")
```

### Aggregate Annual Loss

Having simulated each of the above variables, we use them to construct
an updated expectual annual loss. Notice each simulated variable
captures different components of FEMA's
$\text{EAL} = \text{Exposure} \times AF \times HLR$ formula.

1.  **sim_tornado** - only replaces the AF part for Tornado (measured
    in count);
2.  **sim_flood** - The lognormal distribution handles the
    $HLR\times Exposure$ part for flood hazard for building loss only
    (measured in dollar amount) and Bernoulli distribution handles the
    AF part;
3.  **sim_eq** - same as flood.

Population and agriculture loss components for flood and earthquake,
as well as the full EAL for all non-simulated hazards, are held at
their reconstructed point estimates and added as fixed constants to
every simulated year.

```{r}
#| label: agg-loss

# Tornado: simulated COUNT × fixed (Exposure × HLR)
tornado_dmg_data <- sheldus_county_clean %>%
  filter(Hazard == "Tornado") %>%
  summarise(mean_dmg  = mean(prop_dmg_adj, na.rm = TRUE),
            n_records = n())

cat("Tornado records used for mean damage:", tornado_dmg_data$n_records, "\n")
cat("Fixed mean damage per event (Exposure × HLR per occurrence):",
    dollar(round(tornado_dmg_data$mean_dmg)), "\n")

mean_dmg_per_tornado <- ifelse(
  tornado_dmg_data$n_records > 0,
  tornado_dmg_data$mean_dmg,
  5e6  # if there is no tornado data, use $5M damage as backup
)

tornado_loss <- sim_tornado * mean_dmg_per_tornado

# Flood and earthquake: simulated variable already account for all components of the building loss

# Fixed population & agriculture point estimates, per hazard so sa_run() can use them
eal_tornado_pop_agr <- eal_by_hc %>%
  filter(nri_prefix == "TRND",
         consequence %in% c("Population", "Agriculture")) %>%
  summarise(total = sum(EAL_hc, na.rm = TRUE)) %>%
  pull(total)

eal_flood_pop_agr <- eal_by_hc %>%
  filter(nri_prefix %in% c("RFLD", "CFLD"),
         consequence %in% c("Population", "Agriculture")) %>%
  summarise(total = sum(EAL_hc, na.rm = TRUE)) %>%
  pull(total)

eal_eq_pop_agr <- eal_by_hc %>%
  filter(nri_prefix == "ERQK",
         consequence %in% c("Population", "Agriculture")) %>%
  summarise(total = sum(EAL_hc, na.rm = TRUE)) %>%
  pull(total)

# Reconstructed NRI reference — all consequences, 3 simulated hazards only
# This is the single-point benchmark the MC distribution is compared against
eal_recon_3haz_full <- eal_by_hc %>%
  filter(nri_prefix %in% c("TRND", "RFLD", "CFLD", "ERQK")) %>%
  summarise(eal = sum(EAL_hc, na.rm = TRUE)) %>%
  pull(eal)

# Full EAL for all non-simulated hazards (unchanged)
eal_fixed_other_hazards <- eal_by_hc %>%
  filter(!nri_prefix %in% c("RFLD", "CFLD", "ERQK", "TRND")) %>%
  summarise(total = sum(EAL_hc, na.rm = TRUE)) %>%
  pull(total)

eal_fixed_total <- eal_tornado_pop_agr +
                   eal_flood_pop_agr +
                   eal_eq_pop_agr +
                   eal_fixed_other_hazards

cat("Fixed point estimate breakdown:\n")
cat("  Tornado pop + agr EAL:             ", dollar(round(eal_tornado_pop_agr)), "\n")
cat("  Flood pop + agr EAL:               ", dollar(round(eal_flood_pop_agr)), "\n")
cat("  Earthquake pop + agr EAL:          ", dollar(round(eal_eq_pop_agr)), "\n")
cat("  Full EAL for non-simulated hazards:", dollar(round(eal_fixed_other_hazards)), "\n")
cat("  Total fixed addition per year:     ", dollar(round(eal_fixed_total)), "\n")
cat("  Reconstructed NRI (3 hazards, all consequences):",
    dollar(round(eal_recon_3haz_full)), "\n")

agg_loss <- tornado_loss + sim_flood + sim_eq + eal_fixed_total

#  skewness and kurtosis
agg_mean   <- mean(agg_loss)
agg_sd     <- sd(agg_loss)
skewness_m <- (sum((agg_loss - agg_mean)^3) / N_sim) / agg_sd^3

tibble(
  Statistic = c("Mean ($)", "Median ($)", "SD ($)","95th percentile ($)", "99th percentile ($)","Skewness", "Kurtosis (excess)"),
  Value     = c(dollar(round(mean(agg_loss))),
                dollar(round(median(agg_loss))),
                dollar(round(sd(agg_loss))),
                dollar(round(quantile(agg_loss, 0.95))),
                dollar(round(quantile(agg_loss, 0.99))),
                round(skewness_m, 3),
                round((sum((agg_loss - agg_mean)^4) / N_sim) / agg_sd^4 - 3, 2))
) %>%
  kable(caption = paste0(
  "Summary of Aggregate annual loss")) %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

```{r}
# If we just want to see the simulated part without fixed part
total_sim_loss <- tornado_loss + sim_flood + sim_eq

#  skewness and kurtosis
total_sim_mean   <- mean(total_sim_loss)
total_sim_sd     <- sd(total_sim_loss)
total_sim_skew <- (sum((total_sim_loss - total_sim_mean)^3) / N_sim) / total_sim_sd^3

tibble(
  Statistic = c("Mean ($)", "Median ($)", "SD ($)","95th percentile ($)", "99th percentile ($)","Skewness", "Kurtosis (excess)"),
  Value     = c(dollar(round(mean(total_sim_loss))),
                dollar(round(median(total_sim_loss))),
                dollar(round(sd(total_sim_loss))),
                dollar(round(quantile(total_sim_loss, 0.95))),
                dollar(round(quantile(total_sim_loss, 0.99))),
                round(total_sim_skew, 3),
                round((sum((total_sim_loss - total_sim_mean)^4) / N_sim) / total_sim_sd^4 - 3, 2))
) %>%
  kable(caption = paste0(
  "Summary of annual loss from Tornado, Earthquake, and Flood")) %>%
  kable_styling(bootstrap_options = c("striped","hover"), full_width = FALSE)
```

### Raw Monte Carlo Draws

The exceedance curve that will be shown later is a summary of the
simulation (ranks and sorts the 100,000 draws into a smooth curve).
The plots below show the raw draws themselves, giving us a sense of
what the simulation actually produced year-by-year. Because every
simulated year includes the fixed point-estimate addition of other
un-simulatated variables, no year has zero total loss. TO see the
draws meaningfully, we only looked at the EAL for the three hazards
individually instead of as the total EAL every yeay (because the added
EAL from un-simulated hazards are too big, which makes the graph
unreadable)

```{r}
#| label: mc-hazard-distributions
#| fig-height: 12

# Full-consequence simulated losses 

# Tornado: building (simulated count × fixed mean building damage) + fixed pop and agr point estimates
# Occurrence = any year where Poisson draw > 0
tornado_bldg_loss <- sim_tornado * mean_dmg_per_tornado
tornado_loss_full <- tornado_bldg_loss + eal_tornado_pop_agr
tornado_occurred  <- sim_tornado > 0          # hurdle

# Flood: simulated building loss + fixed pop and agr point estimates
# Occurrence = any year where hurdle-Lognormal draw > 0
flood_loss_full  <- sim_flood + eal_flood_pop_agr
flood_occurred   <- sim_flood > 0             # hurdle

# Earthquake: simulated building loss + fixed pop and agr point estimates
# Occurrence = any year where hurdle-Lognormal draw > 0
eq_loss_full     <- sim_eq + eal_eq_pop_agr
eq_occurred      <- sim_eq > 0               # hurdle

# Occurrence summary 
tibble(
  Hazard = c("Tornado", "Flood", "Earthquake"),
  `Years with occurrence` = comma(c(
    sum(tornado_occurred),
    sum(flood_occurred),
    sum(eq_occurred)
  )),
  `Years without occurrence` = comma(c(
    sum(!tornado_occurred),
    sum(!flood_occurred),
    sum(!eq_occurred)
  )),
  `Pr(occurrence)` = round(c(
    mean(tornado_occurred),
    mean(flood_occurred),
    mean(eq_occurred)
  ), 4),
  `Target Pr(occurrence)` = c(
    round(1 - dpois(0, lambda_T), 4),
    round(p_F_hat, 4),
    p_E
  )
) %>%
  kable(caption = paste0(
    "Simulated occurrence counts across 100,000 annual loss years. ",
    "Simulated probability should closely match target — ",
    "any gap reflects Monte Carlo sampling noise."
  )) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

# Plot (The fixed pop + agr addition is included in every occurrence year)

plot_occurrence_hist <- function(loss_vec, occurred_vec,
                                 hazard_name, fill_color,
                                 fixed_addition) {

  n_occurred  <- sum(occurred_vec)
  n_no_event  <- sum(!occurred_vec)

  # Loss in occurrence years only
  occ_losses <- loss_vec[occurred_vec]

  ggplot(tibble(loss = occ_losses), aes(x = loss)) +
    geom_histogram(bins = 100, fill = fill_color,
                   color = "white", alpha = 0.85) +
    geom_vline(xintercept = mean(occ_losses), linetype = "dashed",
               color = "firebrick", linewidth = 0.9) +
    geom_vline(xintercept = median(occ_losses), linetype = "dotted",
               color = "darkgreen", linewidth = 0.9) +
    annotate("text",
             x = mean(occ_losses) * 1.1, y = Inf,
             label = paste0("Mean (occ. yrs)\n",
                            dollar(round(mean(occ_losses)))),
             color = "firebrick", vjust = 1.5, hjust = 0, size = 3) +
    annotate("text",
             x = median(occ_losses) * 0.9, y = Inf,
             label = paste0("Median (occ. yrs)\n",
                            dollar(round(median(occ_losses)))),
             color = "darkgreen", vjust = 1.5, hjust = 1, size = 3) +
    scale_x_log10(labels = dollar_format(scale = 1e-6, suffix = "M")) +
    scale_y_continuous(labels = comma_format()) +
    labs(
      title    = paste0(hazard_name,
                        " — loss distribution in occurrence years only"),
      subtitle = paste0(
        comma(n_occurred),  " years with hazard occurrence | ",
        comma(n_no_event),  " years with no occurrence (excluded) | ",
        "loss includes building + population + agriculture"
      ),
      x = "Annual loss in occurrence year (log scale, $M)",
      y = "Number of simulated years"
    ) +
    theme_minimal(base_size = 12) +
    theme(plot.title = element_text(face = "bold"))
}

p_tornado <- plot_occurrence_hist(
  tornado_loss_full, tornado_occurred,
  "Tornado",    "goldenrod",
  fixed_addition = eal_tornado_pop_agr
)

p_flood <- plot_occurrence_hist(
  flood_loss_full, flood_occurred,
  "Flood",      "steelblue",
  fixed_addition = eal_flood_pop_agr
)

p_eq <- plot_occurrence_hist(
  eq_loss_full, eq_occurred,
  "Earthquake", "firebrick",
  fixed_addition = eal_eq_pop_agr
)

print(p_tornado / p_flood / p_eq)
```

```{r}
# Only look at nonzero years - otherwise we wont be able to see anything with mostly zero observations in our graphs
df_nonzero <- tibble(loss = total_sim_loss[total_sim_loss > 0])

# splitting points for the first 3/4 observartions (count percentiles)
q25 <- quantile(df_nonzero$loss, 0.25)
q50 <- quantile(df_nonzero$loss, 0.50)
q75 <- quantile(df_nonzero$loss, 0.75)

# Cap the very top at the 99th percentile so the few very year doesn't stretch the axis. The remaining 1/4 are split into 3 range to see tail behaviors (q75, upper_cap] into dollar amount and not counts because our concern is the stretching of the x-axis
upper_cap  <- quantile(df_nonzero$loss, 0.99)
n_excluded <- sum(df_nonzero$loss > upper_cap)
max_loss   <- max(df_nonzero$loss)
tail_breaks <- seq(q75, upper_cap, length.out = 4)
t1 <- tail_breaks[2]
t2 <- tail_breaks[3]

# splitting into the 6 graphs we want
df_p1 <- df_nonzero %>% filter(loss <= q25)                     # 0–25%
df_p2 <- df_nonzero %>% filter(loss > q25 & loss <= q50)        # 25–50%
df_p3 <- df_nonzero %>% filter(loss > q50 & loss <= q75)        # 50–75%
df_p4 <- df_nonzero %>% filter(loss > q75 & loss <= t1)         # tail, 1st third
df_p5 <- df_nonzero %>% filter(loss > t1  & loss <= t2)         # tail, 2nd third
df_p6 <- df_nonzero %>% filter(loss > t2  & loss <= upper_cap)  # tail, 3rd third

# function 
make_panel <- function(df, title, range_label, fill_color, caption = NULL) {
  ggplot(df, aes(x = loss)) +
    geom_histogram(bins = 60, fill = fill_color, color = "white", linewidth = 0.2) +
    scale_x_continuous(labels = label_dollar(scale = 1e-6, suffix = "M"),
                       expand = expansion(mult = c(0, 0.02))) +
    scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.05))) +
    labs(
      title = title,
      subtitle = paste0(comma(nrow(df)), " years · ", range_label),
      caption = caption,
      x = "Annual loss from Tornado, Flood, and Earthquake",
      y = "Simulated years"
    ) +
    theme_minimal(base_size = 12) +
    theme(plot.title = element_text(face = "bold"))
}

# make each of the graphs
p1 <- make_panel(df_p1, "Lowest 25% of non-zero losses",
                 paste0(dollar(round(min(df_nonzero$loss))), " to ", dollar(round(q25))),
                 "#2C5F8D")

p2 <- make_panel(df_p2, "25th–50th percentile",
                 paste0(dollar(round(q25)), " to ", dollar(round(q50))),
                 "#5A7FA8")

p3 <- make_panel(df_p3, "50th–75th percentile",
                 paste0(dollar(round(q50)), " to ", dollar(round(q75))),
                 "#8A5BA6")

p4 <- make_panel(df_p4, "First third of upper quartile (dollar)",
                 paste0(dollar(round(q75)), " to ", dollar(round(t1))),
                 "#B85374")

p5 <- make_panel(df_p5, "Middle third of upper quartile (dollar)",
                 paste0(dollar(round(t1)), " to ", dollar(round(t2))),
                 "#A63A3A")

p6 <- make_panel(df_p6, "Last third of upper quartile (dollar)",
                 paste0(dollar(round(t2)), " to ", dollar(round(upper_cap)),
                        "\n", comma(n_excluded), " extreme yearsexcluded"),
                 "#7A1F1F")

# display graphs
(p1 | p2) / (p3 | p4) / (p5 | p6)


# Summarize exclusion
min_excluded <- min(df_nonzero$loss[df_nonzero$loss > upper_cap])

tibble(
  Metric = c(
    "Total simulated years",
    "Zero-loss years",
    "Non-zero years",
    "Years excluded (above 99th percentile)",
    "Smallest excluded loss",
    "Largest excluded loss"
  ),
  Value = c(
    comma(length(total_sim_loss)),
    comma(sum(total_sim_loss == 0)),
    comma(sum(total_sim_loss > 0)),
    comma(n_excluded),
    dollar(round(min_excluded)),
    dollar(round(max_loss))
  )
) %>%
  kable(
    caption = "Simulated-year accounting and exclusions applied to the histograms",
    align = c("l", "r")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

### Exceedance Probability Curve

This tells us, for any loss level indicated on the x-axis, the
corresponding probability that a given year exceeds that loss. The
further right, the more rare and catastrophic the event. The dashed
line shows the reconstructed NRI point estimate for the same three
hazards across all consequence types, which is the single-number
summary our distribution is meant to replace.

**Note:** We are only looking at the losses from Flood, Tornado, and
earthquake because those are the only part that changed in the NRI
risk calculation, everything else are held at their point-estimate.
This way, we can see more clearly the difference that our replaced
variables made in the risk calculation.

```{r}
# Reconstructed NRI reference — all consequences, 3 simulated hazards (already defined in agg-loss)

# Aggregate loss including fixed pop & agr additions — mirrors sa_run() exactly
agg_loss_full <- tornado_loss_full + flood_loss_full + eq_loss_full
agg_mean_full  <- mean(agg_loss_full)
agg_sd_full    <- sd(agg_loss_full)
skewness_full  <- (sum((agg_loss_full - agg_mean_full)^3) / N_sim) / agg_sd_full^3

loss_sorted <- sort(agg_loss_full)
n_total     <- length(loss_sorted)
surv_prob   <- (n_total:1) / n_total

idx_thin <- round(seq(1, n_total, length.out = 100000))
surv_df  <- tibble(
  loss = loss_sorted[idx_thin],
  surv = surv_prob[idx_thin]
) %>% filter(loss > 0, surv > 0)

#the exceedance probability of where EAL line and exceedance curve intersects
surv_prob_full <- (n_total:1) / n_total
exceed_at_eal  <- surv_prob_full[which.min(abs(loss_sorted - eal_recon_3haz_full))]

ggplot(surv_df, aes(x = loss, y = surv)) +
  geom_line(color = "steelblue", linewidth = 0.75) +
  geom_vline(xintercept = eal_recon_3haz_full, linetype = "dashed",
             color = "firebrick", linewidth = 0.9) +
    geom_hline(yintercept = exceed_at_eal, linetype = "dashed",
             color = "firebrick", linewidth = 0.9) +
  annotate("text",
           x = eal_recon_3haz_full * 1.1, y = 0.25,
           label = paste0("Reconstructed NRI EAL\n(all consequences)\n$",
                          scales::comma(round(eal_recon_3haz_full))),
           color = "firebrick", hjust = 0, size = 3.2) +
  scale_x_log10(labels = dollar_format(scale = 1e-6, suffix = "M")) +
  scale_y_continuous(labels = percent_format(accuracy = 0.01), breaks = sort(unique(c(pretty(surv_df$surv), exceed_at_eal)))) +
  labs(
    title    = "Exceedance Probability Curve",
    subtitle = paste0("Manhattan | Hazards: Tornado, Flood, Earthquake | All consequence types"),
    x        = "Annual aggregate loss (log scale, $M)",
    y        = "Annual exceedance probability"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

### Sensitivity Analysis

**Method:** We use a Local One-at-a-Time (OAT) Sensitivity Analysis
where each of the nine model parameters is individually perturbed to a
plausible range of value while all other parameters are held at their
current estimates. The full 100,000-year Monte Carlo simulation is
re-run for every scenario, and we report three output statistics: mean
aggregate annual loss (including building loss, population loss, and
agriculture loss), 95th-percentile loss, and 99th-percentile loss. We
chose this method because we estimated all 9 parameters independently
(selection of 1 parameter had no impact of any other ones):

1.  $\hat{\lambda}_T$ through Poisson MLE on county-level tornado
    data;
2.  $(\hat{p}_F, \hat{\mu}_F, \hat{\sigma}_F)$ through separate
    Bernoulli and Lognormal MLEs on flood data;
3.  $(p_E, \mu_E, \sigma_E)$ came from published seismic literature.
4.  $(\nu, \rho_{TF})$ chosen to produce heavier joint tails and to
    reflect a weak meteorological link between tornado and flood
    occurrence, respectively.

Due to the fact that these parameters are estimated from distinct data
sources, interaction between parameters are negligible, making this
method both appropriate and easy to implement. We set bounds for
perturbation:

1.  The tornado-rate bounds (×0.5, ×1.5) reflects the sampling
    uncertainty inherent in having only 3 tornado events over 63 years
    of record: $\hat{\lambda}_T = 0.047$, 95% CI: $[-0.006,\ 0.101]$.
2.  The flood Bernoulli and Lognormal bounds are the asymptotic MLE
    95% confidence intervals.\
3.  The earthquake bounds span the full range cited across the
    supporting literature.\

```{r}
# mu_E level (low with signa_E at 0.9 and high at 1.5)
sigma_E_low <- 0.9
sigma_E_high <- 1.5
mu_E_low <- log(4.21e9) - qnorm(0.999) * sigma_E_low
mu_E_high <- log(4.21e9) - qnorm(0.999) * sigma_E_high

```

4.  The copula $\nu$ goes from 3 to $\infty$ (Gaussian Copula) to
    check the transition from heavy joint tails (Student-$t$) to the
    tail-independence of the Gaussian copula.

| Hazard | Parameter | Low | Baseline | High |
|--------------|--------------|--------------|--------------|--------------|
| Tornado | $\lambda_T$ | $\times 0.50$ | MLE | $\times 1.50$ |
| Flood | $\hat{p}_F$ | lower of 95% CI | MLE | upper of 95% CI |
| Flood | $\hat{\mu}_F$ | lower of 95% CI | MLE | upper of 95% CI |
| Flood | $\hat{\sigma}_F$ | lower of 95% CI | MLE | upper of 95% CI |
| Earthquake | $p_E$ | 0.001 | 0.002 | 0.003 |
| Earthquake | $\mu_E$ | `r sigma_E_low` | `r sigma_E` | `r sigma_E_high` |
| Earthquake | $\mu_E$ | `r round(mu_E_low, 2)` | `r round(mu_E, 2)` | `r round(mu_E_high, 2)` |
| Copula | $\nu$ | 3 (heavy) | 4 | 6 / 10 / $\infty$ |
| Copula | $\rho_{TF}$ | 0.00 | 0.05 | 0.15 |

```{r}
#|label: baseline-and-mc-function

# Re-run the full MC for any combination of parameters, this way each scenario is self contained
sa_run <- function(lam_T_in, p_F_in, mu_F_in, sig_F_in,
                   p_E_in,   mu_E_in, sig_E_in,
                   nu_in,    rho_in,
                   seed = 42, N = N_sim) {
  set.seed(seed)
  Ps <- matrix(c(1,      rho_in, 0,
                 rho_in, 1,      0,
                 0,      0,      1), nrow = 3, byrow = TRUE)

  # Handle when nu -> Inf vs a nu as a finite number
  if (is.infinite(nu_in)) {
    U <- pnorm(MASS::mvrnorm(N, mu = rep(0, 3), Sigma = Ps))
  } else {
    U <- pt(rmvt(N, sigma = Ps, df = nu_in), df = nu_in)
  }

  sim_T <- qpois(U[, 1], lambda = lam_T_in)
  sim_F <- as.integer(U[, 2] <= p_F_in)  * qlnorm(runif(N), mu_F_in,  sig_F_in)
  sim_E <- as.integer(U[, 3] <= p_E_in)  * qlnorm(runif(N), mu_E_in,  sig_E_in)
  
# Aggregate: simulated building loss per hazard + fixed pop & agr point estimates
# Mirrors the main simulation in agg-loss exactly
  agg <- sim_T * mean_dmg_per_tornado + eal_tornado_pop_agr +
       sim_F + eal_flood_pop_agr +
       sim_E + eal_eq_pop_agr

  tibble(
    mean_loss = mean(agg),
    p95_loss  = as.numeric(quantile(agg, 0.95)),
    p99_loss  = as.numeric(quantile(agg, 0.99))
  )
}

# Baseline (our original estimated parameter)
sa_base <- sa_run(
  lam_T_in = lambda_T, p_F_in   = p_F_hat, mu_F_in  = mu_F,    sig_F_in = sigma_F,
  p_E_in   = p_E,      mu_E_in  = mu_E,    sig_E_in = sigma_E,
  nu_in    = nu,       rho_in   = 0.05
)

cat("Sensitivity Analysis baseline — Mean:", dollar(round(sa_base$mean_loss)),
    "| P95:", dollar(round(sa_base$p95_loss)),
    "| P99:", dollar(round(sa_base$p99_loss)), "\n")
```

**Note:** Notice the baseline mean us higher than the 95th percentile.
This is not a bug, but rather it is the result of a aggregate
distribution with fat tail. Because in our data, most years have
near-zero loss but some years experience catastrophic event. The tail
event will drag the mean above the 95th percentile.

```{r}
#|label: CI-bound-for-Flood

# Bernoulli: using the 95% we got from section 2.4
p_F_lo <- max(0, p_F_hat - 1.96 * p_F_se)
p_F_hi <- min(1, p_F_hat + 1.96 * p_F_se)

# Lognormal
n_fl <- n_flood_loss_years

mu_F_lo  <- mu_F    - 1.96 * flood_fit$sd["meanlog"]
mu_F_hi  <- mu_F    + 1.96 * flood_fit$sd["meanlog"]
sig_F_lo <- sigma_F - 1.96 * flood_fit$sd["sdlog"]
sig_F_hi <- sigma_F + 1.96 * flood_fit$sd["sdlog"]

tibble(
  Bound          = c("p_F low (95% CI)", "p_F high (95% CI)",
                     "mu_F low (95% CI)", "mu_F high (95% CI)",
                     "sigma_F low (95% CI)", "sigma_F high (95% CI)"),
  Value          = round(c(p_F_lo, p_F_hi, mu_F_lo, mu_F_hi, sig_F_lo, sig_F_hi), 4),
  `Baseline`     = round(c(p_F_hat, p_F_hat, mu_F, mu_F, sigma_F, sigma_F), 4)
) %>%
  kable(caption = "Flood parameters perturbation bounds — MLE 95% confidence intervals") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

Then the updated OAT table is:

| Hazard | Parameter | Low | Baseline | High |
|--------------|--------------|-----------------|-----------|----------------|
| Tornado | $\lambda_T$ | $\times 0.50$ | MLE | $\times 1.50$ |
| Flood | $\hat{p}_F$ | lower of 95% CI: `r round(p_F_lo,2)` | MLE | upper of 95% CI: `r round(p_F_hi,2)` |
| Flood | $\hat{\mu}_F$ | lower of 95% CI: `r round(mu_F_lo,2)` | MLE | upper of 95% CI: `r round(mu_F_hi,2)` |
| Flood | $\hat{\sigma}_F$ | lower of 95% CI: `r round(sig_F_lo,2)` | MLE | upper of 95% CI: `r round(sig_F_hi,2)` |
| Earthquake | $p_E$ | 0.001 | 0.002 | 0.003 |
| Earthquake | $\mu_E$ | `r sigma_E_low` | `r sigma_E` | `r sigma_E_high` |
| Earthquake | $\mu_E$ | `r round(mu_E_low, 2)` | `r round(mu_E, 2)` | `r round(mu_E_high, 2)` |
| Copula | $\nu$ | 3 (heavy) | 4 | 6 / 10 / $\infty$ |
| Copula | $\rho_{TF}$ | 0.00 | 0.05 | 0.15 |

```{r}
#|label: scenario

# Each row is one scenario of sensitivity analysis.

sa_grid <- tribble(
  ~group,        ~param_id,    ~direction, ~label,
  ~lam_T_v,          ~p_F_v,   ~mu_F_v, ~sig_F_v,
  ~p_E_v,            ~mu_E_v,  ~sig_E_v,
  ~nu_v,             ~rho_v,

# Baseline
  "Baseline",    "baseline",   "base",    "Baseline",
  lambda_T,          p_F_hat,  mu_F,    sigma_F,
  p_E,               mu_E,     sigma_E,
  nu,                0.05,

# Tornado: lambda_T
  "Tornado",     "lambda_T",   "low",     "lambda_T x0.50  (sparse-data lower)",
  lambda_T*0.5,      p_F_hat,  mu_F,    sigma_F,
  p_E,               mu_E,     sigma_E,
  nu,                0.05,

  "Tornado",     "lambda_T",   "high",    "lambda_T x1.50  (sparse-data upper)",
  lambda_T*1.5,      p_F_hat,  mu_F,    sigma_F,
  p_E,               mu_E,     sigma_E,
  nu,                0.05,

# Flood: p_F
  "Flood",       "p_F",        "low",     "p_F  95% CI lower",
  lambda_T,          p_F_lo,   mu_F,    sigma_F,
  p_E,               mu_E,     sigma_E,
  nu,                0.05,

  "Flood",       "p_F",        "high",    "p_F  95% CI upper",
  lambda_T,          p_F_hi,   mu_F,    sigma_F,
  p_E,               mu_E,     sigma_E,
  nu,                0.05,

# Flood: mu_F
  "Flood",       "mu_F",       "low",     "mu_F  MLE 95% CI lower",
  lambda_T,          p_F_hat,  mu_F_lo, sigma_F,
  p_E,               mu_E,     sigma_E,
  nu,                0.05,

  "Flood",       "mu_F",       "high",    "mu_F  MLE 95% CI upper",
  lambda_T,          p_F_hat,  mu_F_hi, sigma_F,
  p_E,               mu_E,     sigma_E,
  nu,                0.05,

# Flood: sigma_F 
  "Flood",       "sigma_F",    "low",     "sigma_F  MLE 95% CI lower",
  lambda_T,          p_F_hat,  mu_F,    sig_F_lo,
  p_E,               mu_E,     sigma_E,
  nu,                0.05,

  "Flood",       "sigma_F",    "high",    "sigma_F  MLE 95% CI upper",
  lambda_T,          p_F_hat,  mu_F,    sig_F_hi,
  p_E,               mu_E,     sigma_E,
  nu,                0.05,

# Earthquake: p_E
  "Earthquake",  "p_E",        "low",     "p_E = 0.001  (literature lower)",
  lambda_T,          p_F_hat,  mu_F,    sigma_F,
  0.001,             mu_E,     sigma_E,
  nu,                0.05,
  
"Earthquake",  "p_E",        "high",    "p_E = 0.003  (literature upper)",
  lambda_T,          p_F_hat,  mu_F,    sigma_F,
  0.003,             mu_E,     sigma_E,
  nu,                0.05,

# Earthquake: mu_E
  "Earthquake",  "mu_E", "low",  paste0("sigma_E = ", sigma_E_low, " → mu_E = ", round(mu_E_low, 2)),
  lambda_T, p_F_hat, mu_F, sigma_F,
  p_E, mu_E_low, sigma_E_low,
  nu, 0.05,


  "Earthquake",  "mu_E", "high", paste0("sigma_E = ", sigma_E_high, " → mu_E = ", round(mu_E_high, 2)),
  lambda_T, p_F_hat, mu_F, sigma_F,
  p_E, mu_E_low, sigma_E_high,
  nu, 0.05,

# Earthquake: sigma_E
  "Earthquake",  "sigma_E",    "low",     "sigma_E = 0.9",
  lambda_T,          p_F_hat,  mu_F,    sigma_F,
  p_E,               mu_E,     0.7,
  nu,                0.05,

  "Earthquake",  "sigma_E",    "high",    "sigma_E = 1.5 ",
  lambda_T,          p_F_hat,  mu_F,    sigma_F,
  p_E,               mu_E,     1.3,
  nu,                0.05,

# Copula: nu
  "Copula",      "nu",         "low",     "nu = 3  (heaviest joint tail)",
  lambda_T,          p_F_hat,  mu_F,    sigma_F,
  p_E,               mu_E,     sigma_E,
  3,                 0.05,

  "Copula",      "nu",         "high_6",  "nu = 6",
  lambda_T,          p_F_hat,  mu_F,    sigma_F,
  p_E,               mu_E,     sigma_E,
  6,                 0.05,

  "Copula",      "nu",         "high_10", "nu = 10",
  lambda_T,          p_F_hat,  mu_F,    sigma_F,
  p_E,               mu_E,     sigma_E,
  10,                0.05,

  "Copula",      "nu",         "high",    "nu = Inf  (Gaussian copula)",
  lambda_T,          p_F_hat,  mu_F,    sigma_F,
  p_E,               mu_E,     sigma_E,
  Inf,               0.05,

# Copula: rho_TF
  "Copula",      "rho_TF",     "low",     "rho_TF = 0.00  (independence)",
  lambda_T,          p_F_hat,  mu_F,    sigma_F,
  p_E,               mu_E,     sigma_E,
  nu,                0.00,

  "Copula",      "rho_TF",     "high",    "rho_TF = 0.15  (moderate weather link)",
  lambda_T,          p_F_hat,  mu_F,    sigma_F,
  p_E,               mu_E,     sigma_E,
  nu,                0.15
)
```

```{r}
# Running all scenarios
sa_results <- sa_grid %>%
  mutate(
    res = pmap(
      list(lam_T_v, p_F_v, mu_F_v, sig_F_v,
           p_E_v,   mu_E_v, sig_E_v,
           nu_v,    rho_v),
      function(a, b, c, d, e, f, g, h, i)
        sa_run(lam_T_in = a, p_F_in   = b, mu_F_in  = c, sig_F_in = d,
               p_E_in   = e, mu_E_in  = f, sig_E_in = g,
               nu_in    = h, rho_in   = i)
    )
  ) %>%
  unnest(res) %>%
  mutate(
    chg_mean = (mean_loss - sa_base$mean_loss) / sa_base$mean_loss * 100,
    chg_p95  = (p95_loss  - sa_base$p95_loss)  / sa_base$p95_loss  * 100,
    chg_p99  = (p99_loss  - sa_base$p99_loss)  / sa_base$p99_loss  * 100
  )
```

```{r}

# Sensitivity table for each hazard group
# Reports mean, P95, and P99 of aggregate annual loss (building + population + agriculture
sa_table <- function(grp, caption_text) {
  sa_results %>%
    filter(group %in% c("Baseline", grp)) %>%
    transmute(
      Scenario        = label,
      `Mean Loss`     = dollar(round(mean_loss)),
      `Delta Mean`    = ifelse(direction == "base", "—",
                               sprintf("%+.3f%%", chg_mean)),
      `95th Pct`      = dollar(round(p95_loss)),
      `Delta P95`     = ifelse(direction == "base", "—",
                               sprintf("%+.3f%%", chg_p95)),
      `99th Pct`      = dollar(round(p99_loss)),
      `Delta P99`     = ifelse(direction == "base", "—",
                               sprintf("%+.3f%%", chg_p99))
    ) %>%
    kable(caption = caption_text, align = c("l","r","r","r","r","r","r")) %>%
    kable_styling(bootstrap_options = c("striped", "hover"),
                  full_width = FALSE) %>%
    row_spec(1, bold = TRUE, background = "#f5f5f5")   # highlight baseline row
}
```

#### Tornado: Sensitivity to Annual Frequency Rate ($\lambda_T$)

The Poisson rate $\lambda_T = 0.047$ was estimated from only 3 tornado
events in 63 years of record, giving a wide 95% CI of
$[-0.006, 0.101]$. We updated this parameter by $\pm 50\%$ because
tornado damage per event is small (FEMA reported an average of
\$22,681), so even a $150\%$ rate result in a mild effect on aggregate
loss.

```{r}
#|label: Tornado-sa
sa_table("Tornado",
         "Tornado sensitivity")
```

#### Flood: Sensitivity to Bernoulli (Occurrence) and Lognormal (Severity)

We tested all three flood parameters, which are the Bernoulli
probability $\hat{p}_F$ and the Lognormal parameters
$(\hat{\mu}_F, \hat{\sigma}_F)$. We shifted each of them to the 95% CI
of MLE. Because the flood distribution has a fat right tail, the
severity parameters ($\mu_F$, $\sigma_F$) tend to have a
disproportionately large effect on the 99th-percentile loss relative.

```{r}
#|label: Flood-sa
sa_table("Flood",
         "Flood sensitivity")
```

#### Earthquake: Sensitivity to Bernoulli (Occurrence) and Lognormal (Severity)

We derived all earthquake parameters from external literature rather
than fitting them using historical data. The occurrence probability
$p_E$ spans the $[0.001, 0.003]$ range as backed by literature. The
severity parameters $\mu_E$ and $\sigma_E$ are shifted by $\pm 0.3$
log-units and $\pm 0.3$, respectively.

```{r}
#|label: Earthquake-sa
sa_table("Earthquake",
         "Earthquake sensitivity")
```

#### Sensitivity of 99th-Percentile Aggregate Loss

The tornado diagram (name of the diagram, not the hazard) below ranks
all parameters by the total range of change they produce in the
99th-percentile of aggregate loss. This diagram answers the question
of: "which parameter, if we are wrong about it, threatens our estimate
of catastrophic loss the most?" Each bar shows the change from
baseline when the parameter is set to its low (left, negative) or high
(right, positive) value. The parameter whose bars extend furthest from
center dominate the uncertainty in tail of our risk distribution.

```{r}
#| label: tornado-diagram
#| fig-height: 6
#| fig-width: 10

param_label_plain <- c(
  lambda_T = "λ_T  (Tornado rate)",
  p_F      = "p̂_F  (Flood occurrence)",
  mu_F     = "μ̂_F  (Flood log-mean)",
  sigma_F  = "σ̂_F  (Flood log-SD)",
  p_E      = "p_E  (EQ occurrence)",
  mu_E     = "μ_E  (EQ log-mean)",
  sigma_E  = "σ_E  (EQ log-SD)",
  nu       = "ν    (Copula d.f.)",
  rho_TF   = "ρ_TF (Tornado–Flood corr.)"
)

tornado_df <- sa_results %>%
  filter(group != "Baseline") %>%
  group_by(group, param_id) %>%
  summarise(
    lo_p99 = min(chg_p99),
    hi_p99 = max(chg_p99),
    lo_p95 = min(chg_p95),
    hi_p95 = max(chg_p95),
    .groups = "drop"
  ) %>%
  mutate(range_p99 = hi_p99 - lo_p99) %>%
  arrange(range_p99) %>%                            # sort ascending by range
  mutate(
    param_label = dplyr::recode(param_id,           # recode AFTER arrange
      lambda_T = "λ_T  (Tornado rate)",
      p_F      = "p̂_F  (Flood occurrence)",
      mu_F     = "μ̂_F  (Flood log-mean)",
      sigma_F  = "σ̂_F  (Flood log-SD)",
      p_E      = "p_E  (EQ occurrence)",
      mu_E     = "μ_E  (EQ log-mean)",
      sigma_E  = "σ_E  (EQ log-SD)",
      nu       = "ν    (Copula d.f.)",
      rho_TF   = "ρ_TF (Tornado–Flood corr.)"
    ),
    y_label = factor(param_label, levels = param_label)  # freeze sort order
  )

group_colours <- c(
  Tornado    = "goldenrod",
  Flood      = "steelblue",
  Earthquake = "firebrick",
  Copula     = "mediumpurple"
)

ggplot(tornado_df) +
  geom_vline(xintercept = 0, colour = "grey40", linewidth = 0.6) +
  geom_segment(aes(x = lo_p99, xend = 0,
                   y = y_label, yend = y_label,
                   colour = group),
               linewidth = 7, lineend = "butt", alpha = 0.85) +
  geom_segment(aes(x = 0, xend = hi_p99,
                   y = y_label, yend = y_label,
                   colour = group),
               linewidth = 7, lineend = "butt", alpha = 0.85) +
  geom_text(aes(x = lo_p99 - 0.5, y = y_label,
                label = sprintf("%.1f%%", lo_p99)),
            hjust = 1, size = 3, colour = "grey30") +
  geom_text(aes(x = hi_p99 + 0.5, y = y_label,
                label = sprintf("+%.1f%%", hi_p99)),
            hjust = 0, size = 3, colour = "grey30") +
  scale_colour_manual(values = group_colours, name = "Hazard group") +
  scale_x_continuous(
    labels = function(x) paste0(ifelse(x >= 0, "+", ""), round(x), "%"),
    expand = expansion(mult = c(0.18, 0.18))
  ) +
  labs(
    title    = "Sensitivity of 99th-Percentile Aggregate Loss",
    subtitle = paste0(
      "Each bar shows the % change in P99 when a single parameter is moved to its low (left) or high (right) bound.\n",
      "Parameters sorted by total range (widest = most influential)."
    ),
    x       = "% change in 99th-percentile aggregate annual loss (building + population + agriculture)",
    y       = NULL,
    caption = paste0(
      "Reconstructed NRI reference — TRND + RFLD + CFLD + ERQK, all consequences: ",
      dollar(round(eal_recon_3haz_full))
    )
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title         = element_text(face = "bold"),
    plot.subtitle      = element_text(size = 9, colour = "grey40"),
    legend.position    = "bottom",
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank()
  )
```

#### Copula's Sensitivity (To Tail-Dependence Strength ($\nu$) and Tornado–Flood Correlation ($\rho_{TF}$))

The degrees-of-freedom parameter $\nu$ of t-Copula tells us how much
joint tail probability are placed on scenarios where multiple hazards
are extreme simultaneously. As $\nu \to \infty$ the t-Copula converges
to the Gaussian copula, which has zero tail dependence. We range
$\nu \in \{3, 4, 6, 10, \infty\}$ to check how strongly our tail-risk
estimates depend on our choice for this parameter. The correlation
$\rho_{TF}$ ranges from full independence (0) to a moderate
weather-driven link (0.15).

```{r}
sa_table("Copula",
         "Copula sensitivity")
```

The degrees-of-freedom parameter $\nu$ is a choice rather than an
estimated scalar.

```{r}
nu_sweep_df <- sa_results %>%
  filter(param_id == "nu" | direction == "base") %>%
  # Recover the nu value for each row
  mutate(nu_val = case_when(
    direction == "base"     ~ as.numeric(nu),   # baseline nu (from env)
    grepl("nu = 3",  label) ~ 3,
    grepl("nu = 4",  label) ~ 4,
    grepl("nu = 6",  label) ~ 6,
    grepl("nu = 10", label) ~ 10,
    grepl("Gaussian",label) ~ Inf,
    TRUE                    ~ NA_real_
  )) %>%
  filter(!is.na(nu_val)) %>%
  # For Inf, use a large finite value so the x-axis looks clean
  mutate(nu_plot = ifelse(is.infinite(nu_val), 40, nu_val)) %>%
  select(nu_val, nu_plot, mean_loss, p95_loss, p99_loss) %>%
  pivot_longer(cols = c(mean_loss, p95_loss, p99_loss),
               names_to = "stat", values_to = "loss") %>%
  mutate(stat = recode(stat,
                       mean_loss = "Mean",
                       p95_loss  = "95th percentile",
                       p99_loss  = "99th percentile"))

p_nu_sweep <- ggplot(nu_sweep_df, aes(x = nu_plot, y = loss / 1e6,
                                       colour = stat, group = stat)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_x_continuous(
    breaks = c(3, 4, 6, 10, 40),
    labels = c("3", "4", "6", "10", "Gaussian\n(∞)")
  ) +
  scale_y_continuous(labels = dollar_format(suffix = "M", scale = 1)) +
  scale_colour_manual(
    values = c("Mean" = "grey50",
               "95th percentile" = "steelblue",
               "99th percentile" = "firebrick"),
    name = NULL
  ) +
  labs(
    title    = "Copula Degrees-of-Freedom and Its Effect on Aggregate Loss Statistics",
    x        = "Copula degrees of freedom (nu)",
    y        = "Aggregate annual loss — all consequences ($M)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold"),
    legend.position = "bottom"
  )

print(p_nu_sweep)
```

# Conclusion

The main goal of my project is to make the point that FEMA's National
Risk Index, by reporting a single mean-based Expected Annual Loss,
hides more than it reveals about the true risk faced by our
communities. The reconstruction of the NRI for New York County
(Manhattan) confirmed that the EAL formula — Exposure × Annualized
Frequency × Historic Loss Ratio — can be closely replicated. Although
our composite EAL of $\$476M$ is about $155.3%$ of FEMA's published
$\$306.7M$, which is a discrepancy attributable to differences in
SHELDUS date ranges and FEMA's proprietary pre-processing. Our
Coefficient of Variation (CV) analysis showed that for most hazards at
the county scale, the mean loss ratio seems to be dominated by a
handful of extreme events rather than the typical year. Having tail
events like this is precisely why I wanted to highlight the flaw in
collapsing the entire distribution into a single number. Because with
just one number, we have no room to see that most years experience
none to minimal hazard losses but a few experiences catastrophic
losses.

This point is illustrated using a Monte Carlo simulation. Across our
100,000 simulated years, the aggregate annual loss across all
consequence types for the three simulated hazards had a mean of
`r dollar(round(mean(agg_loss_full)))` with a standard deviation of
`r dollar(round(sd(agg_loss_full)))` and skewness of
`r round(skewness_m, 0)` We also see that our distribution so
right-skewed that the mean sits above the 95th percentile. Our
exceedance probability curve showed what FEMA's single estimate wasn't
able to show, which is that there is a probable chance of a year with
near-zero loss, but also a non-trivial amount of scenarios where a
major flood or earthquake produces losses in the hundreds of millions
or more. Then our sensitivity analysis further showed that the flood
log-severity parameter (σ̂\_F) is the most influential source of
uncertainty in the tail, with its upper 95% CI bound increasing the
99th-percentile loss by over 400%, while the copula's
degrees-of-freedom parameter confirmed that our choice of a t-Copula
instead of a Gaussian copula was warranted because the Gaussian
copula's lack of ability to account for tail independence.

The above findings taken together suggest that any risk communication
framework that reports only a mean EAL without an accompanying
distribution is probably inadequate for important decisions that
depend on tail outcomes. For example, infrastructure investment,
insurance pricing, emergency reserve planning, and long-term fiscal
resilience. For a county like Manhattan, where building exposure per
square mile is the highest in New York State and most of the United
States, a rare but plausible earthquake or major flood event can lead
to losses way above the annual average computed in FEMA's NRI.
